---
title: "L2 variation diagnostic — where's the action?"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    highlight: kate
    theme: flatly
---

Standalone diagnostic. Self-contained load — does not touch
`refusal_analysis.Rmd`. Goal: locate where L2 form features vary most.
Three lenses:

1. **Across-harm dispersion by model** — does model M use the same L2 forms regardless of harm type, or does it adapt?
2. **Across-model dispersion by harm** — for harm type H, do models converge on a common refusal form, or diverge?
3. **Per-feature dispersion** — which of the 13 L2 features carries the most signal across the (model × harm) grid?

Dispersion measure: SD of cell rates across the relevant axis, then averaged
across L2 features. (For binary features, SD of `p` across cells captures
between-cell spread; variance and Bernoulli entropy give the same ranking.)
Cells with `n_nc < MIN_N` are filtered out to keep small-cell noise from
dominating.

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6, dpi = 110)
library(tidyverse)
library(scales)
library(knitr)
library(kableExtra)

# ---- Load (mirrors refusal_analysis.Rmd setup) ------------------------------
ANNOTATION_DIR <- file.path("..", "data", "annotations")
PROMPT_SHA256  <- "238bb923037bb4126bef041f202fd6547e328aa62f4f61f06b8f63d784891fb1"

L2_FEATURES <- c(
  "explicit_nc", "implicit_nc", "apology", "hedge", "explanatory_preface",
  "positive_alignment", "solidarity", "negative_stance",
  "executed_alternative", "alternative_offer", "normative_suggestion",
  "statement_of_principle", "role_based"
)

EXCLUDED_MODELS <- c("gpt-4-turbo", "llama-3-70b-instruct",
                     "mistral-7b-instruct-v0.1")

read_judged <- function(path) {
  model <- str_replace(basename(path), "^responses_(.*)\\.judged\\.csv$", "\\1")
  suppressWarnings(readr::read_csv(path, show_col_types = FALSE,
    col_types = cols(.default = col_character()))) |>
    mutate(model = model)
}

to_int_binary <- function(x) {
  x <- na_if(trimws(x), "")
  suppressWarnings(as.integer(x))
}

raw <- list.files(ANNOTATION_DIR, pattern = "^responses_.*\\.judged\\.csv$",
                  full.names = TRUE) |>
  map_dfr(read_judged) |>
  filter(!model %in% EXCLUDED_MODELS)

if (!"judge_skipped" %in% names(raw)) raw$judge_skipped <- NA_character_

df <- raw |>
  mutate(across(all_of(L2_FEATURES), to_int_binary),
         parse_error   = if_else(parse_error   %in% c("True","TRUE","1"),
                                 TRUE, FALSE, missing = FALSE),
         judge_skipped = if_else(judge_skipped %in% c("True","TRUE","1"),
                                 TRUE, FALSE, missing = FALSE)) |>
  filter(is.na(judge_prompt_sha256) | judge_prompt_sha256 == "" |
           judge_prompt_sha256 == PROMPT_SHA256)

df_nc <- df |>
  filter(!parse_error, !judge_skipped, layer0 == "Non-compliance")

cat("In-voice NC rows used:", nrow(df_nc), "\n",
    "Models:", n_distinct(df_nc$model), "\n",
    "Harm categories:", n_distinct(df_nc$primary_llama_guard_category_name),
    "\n")
```

```{r grid-setup}
# Minimum cell size to include in dispersion calcs. Cells smaller than this
# get filtered: their per-feature rates are too noisy to inform a SD.
MIN_N <- 5

# Long: one row per (model, harm, feature) with rate + n
cell_long <- df_nc |>
  filter(!is.na(primary_llama_guard_category_name)) |>
  group_by(model, harm = primary_llama_guard_category_name) |>
  summarise(
    n_nc = n(),
    across(all_of(L2_FEATURES), \(x) mean(x, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  pivot_longer(all_of(L2_FEATURES), names_to = "feature", values_to = "rate") |>
  filter(n_nc >= MIN_N)

cat("(model × harm) cells retained at MIN_N =", MIN_N, ":",
    n_distinct(paste(cell_long$model, cell_long$harm)),
    "of",
    n_distinct(df_nc$model) * n_distinct(df_nc$primary_llama_guard_category_name),
    "\n")
```

---

# 1. Across-harm dispersion by model

For each model, take its (harm × feature) rate matrix. Within each feature
column, compute SD across harm categories — this captures how much the
feature's rate moves as the harm type changes. Average across features to get
a single per-model "harm-sensitivity" score: higher = the model's L2 output
shifts more across harm categories; lower = the model uses roughly the same
form everywhere.

```{r across-harm-by-model}
across_harm_by_model <- cell_long |>
  group_by(model, feature) |>
  summarise(sd_across_harm = sd(rate, na.rm = TRUE), .groups = "drop") |>
  group_by(model) |>
  summarise(mean_sd = mean(sd_across_harm, na.rm = TRUE),
            .groups = "drop") |>
  arrange(desc(mean_sd))

across_harm_by_model |>
  mutate(model = fct_reorder(model, mean_sd)) |>
  ggplot(aes(mean_sd, model)) +
  geom_col(fill = "steelblue") +
  scale_x_continuous(labels = number_format(accuracy = 0.001)) +
  labs(x = "Mean SD of L2 rate across harm categories",
       y = NULL,
       title = "How much does each model's L2 form shift across harm types?",
       subtitle = "Higher = more harm-adaptive; lower = more uniform refusal style") +
  theme_minimal(base_size = 11)

across_harm_by_model |>
  mutate(mean_sd = round(mean_sd, 4)) |>
  kable(caption = "Per-model harm-sensitivity: mean (over L2 features) of SD (over harm categories) of feature rates.") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## 1a. Per-model × per-feature heatmap

Same data, unaggregated: shows *which features* are responsible for each
model's harm-sensitivity score.

```{r across-harm-by-model-feature, fig.height = 8}
heat_mf <- cell_long |>
  group_by(model, feature) |>
  summarise(sd_across_harm = sd(rate, na.rm = TRUE), .groups = "drop")

# Order features by total dispersion, models by overall harm-sensitivity
feat_order  <- heat_mf |> group_by(feature) |>
  summarise(s = sum(sd_across_harm, na.rm = TRUE)) |> arrange(s) |> pull(feature)
model_order <- across_harm_by_model |> arrange(mean_sd) |> pull(model)

heat_mf |>
  mutate(feature = factor(feature, levels = feat_order),
         model   = factor(model,   levels = model_order)) |>
  ggplot(aes(feature, model, fill = sd_across_harm)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(option = "C", labels = number_format(accuracy = 0.001)) +
  labs(x = NULL, y = NULL, fill = "SD across\nharm cats",
       title = "Per (model × feature) SD of rate across harm categories") +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

# 2. Across-model dispersion by harm

Flip the axis. For each harm category, take its (model × feature) rate
matrix. Within each feature column, compute SD across models — this captures
how much models disagree on the appropriate L2 form for this kind of harm.
Average across features for a per-harm "controversy" score.

```{r across-model-by-harm}
across_model_by_harm <- cell_long |>
  group_by(harm, feature) |>
  summarise(sd_across_model = sd(rate, na.rm = TRUE), .groups = "drop") |>
  group_by(harm) |>
  summarise(mean_sd = mean(sd_across_model, na.rm = TRUE),
            .groups = "drop") |>
  arrange(desc(mean_sd))

across_model_by_harm |>
  mutate(harm = fct_reorder(harm, mean_sd)) |>
  ggplot(aes(mean_sd, harm)) +
  geom_col(fill = "tomato3") +
  scale_x_continuous(labels = number_format(accuracy = 0.001)) +
  labs(x = "Mean SD of L2 rate across models",
       y = NULL,
       title = "Which harm categories elicit the most divergent L2 forms across models?",
       subtitle = "Higher = models disagree more on the refusal form for this harm type") +
  theme_minimal(base_size = 11)

across_model_by_harm |>
  mutate(mean_sd = round(mean_sd, 4)) |>
  kable(caption = "Per-harm-category controversy: mean (over L2 features) of SD (over models) of feature rates.") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## 2a. Per-harm × per-feature heatmap

```{r across-model-by-harm-feature, fig.height = 7}
heat_hf <- cell_long |>
  group_by(harm, feature) |>
  summarise(sd_across_model = sd(rate, na.rm = TRUE), .groups = "drop")

feat_order_h <- heat_hf |> group_by(feature) |>
  summarise(s = sum(sd_across_model, na.rm = TRUE)) |> arrange(s) |> pull(feature)
harm_order   <- across_model_by_harm |> arrange(mean_sd) |> pull(harm)

heat_hf |>
  mutate(feature = factor(feature, levels = feat_order_h),
         harm    = factor(harm,    levels = harm_order)) |>
  ggplot(aes(feature, harm, fill = sd_across_model)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(option = "C", labels = number_format(accuracy = 0.001)) +
  labs(x = NULL, y = NULL, fill = "SD across\nmodels",
       title = "Per (harm × feature) SD of rate across models") +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

# 3. Per-feature dispersion across the full (model × harm) grid

Which L2 features carry the most signal full-stop? For each feature,
compute SD of its rate across every (model × harm) cell that passes the
MIN_N filter. Also report the cell mean rate so we can read floor/ceiling
features (e.g. explicit_nc near 1 everywhere) against actually-bimodal ones.

```{r per-feature}
per_feature <- cell_long |>
  group_by(feature) |>
  summarise(
    mean_rate = mean(rate, na.rm = TRUE),
    sd_cells  = sd(rate, na.rm = TRUE),
    min_rate  = min(rate, na.rm = TRUE),
    max_rate  = max(rate, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(sd_cells))

per_feature |>
  mutate(feature = fct_reorder(feature, sd_cells)) |>
  ggplot(aes(sd_cells, feature)) +
  geom_col(fill = "darkorchid4") +
  geom_point(aes(x = mean_rate), color = "grey30", size = 2) +
  scale_x_continuous(labels = number_format(accuracy = 0.01),
                     sec.axis = sec_axis(~., name = "● mean cell rate",
                                         labels = number_format(accuracy = 0.01))) +
  labs(x = "SD of feature rate across (model × harm) cells",
       y = NULL,
       title = "Where's the L2 signal? Per-feature dispersion across the grid",
       subtitle = "Bar = SD across cells (between-cell spread); dot = mean cell rate") +
  theme_minimal(base_size = 11)

per_feature |>
  mutate(across(c(mean_rate, sd_cells, min_rate, max_rate),
                \(x) percent(x, accuracy = 0.1))) |>
  kable(caption = "Per-feature dispersion across (model × harm) cells. Read sd_cells against mean_rate: a low SD on a near-floor/ceiling feature is unsurprising; a low SD on a mid-range mean means real convergence.") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# 4. Joint reader: top-variation features in context

For the top-3 most dispersive features (from §3), show their full
(model × harm) rate matrix as a heatmap. This is where the action is.

```{r top-feature-heatmaps, fig.height = 5}
top_feats <- per_feature |> slice_max(sd_cells, n = 3) |> pull(feature)

for (f in top_feats) {
  p <- cell_long |>
    filter(feature == f) |>
    ggplot(aes(harm, model, fill = rate)) +
    geom_tile(color = "white") +
    scale_fill_viridis_c(option = "B", limits = c(0, 1),
                         labels = percent_format(accuracy = 1)) +
    labs(x = NULL, y = NULL, fill = "rate",
         title = paste0("Feature: ", f, " — rate by (model × harm)")) +
    theme_minimal(base_size = 10) +
    theme(axis.text.x = element_text(angle = 35, hjust = 1))
  print(p)
}
```

---

# Notes

- Denominator throughout is **in-voice NC only** (service-level rows and
  parse-errors filtered out). A model that rarely refuses for some harm
  category gets a smaller-n cell that may be filtered by `MIN_N`.
- All-zero columns within a (model × harm) cell yield 0 rate — distinct from
  NA. The SD machinery treats those as real zeros, which is correct for L2
  ("model never used apology on these prompts").
- Variance vs Bernoulli entropy: for binary features, both rank cells the
  same way (entropy is monotone in `p(1-p)` for `p ≤ 0.5`); SD is more
  interpretable as "percentage-point spread".
