---
title: "LLM Refusal Strategy — Descriptive Analysis"
author: "Polite Refusals team"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 6, dpi = 110
)
library(tidyverse)
library(jsonlite)
library(scales)
library(knitr)
library(kableExtra)

# ---- Config -----------------------------------------------------------------
ANNOTATION_DIR <- file.path("..", "data", "annotations")
# Current judge prompt SHA256 (from BASE_SYSTEM_PROMPT in scripts/utils/judge_config.py).
# Records whose judge_prompt_sha256 != this are dropped, defending against
# accidental mixing of runs after future prompt revisions.
PROMPT_SHA256 <- "238bb923037bb4126bef041f202fd6547e328aa62f4f61f06b8f63d784891fb1"

L2_FEATURES <- c(
  "explicit_nc", "implicit_nc", "apology", "hedge", "explanatory_preface",
  "positive_alignment", "solidarity", "negative_stance",
  "executed_alternative", "alternative_offer", "normative_suggestion",
  "statement_of_principle", "role_based"
)

# Model display order: group by family, with the non-reasoning sibling first.
MODEL_ORDER <- c(
  # OpenAI
  "gpt-4o", "gpt-5", "gpt-5.3", "gpt-5.3-reasoning",
  # Anthropic
  "claude-sonnet-3.7", "claude-sonnet-4.6", "claude-opus-4.6", "claude-opus-4.6-reasoning",
  # Google
  "gemini-2.5-pro",
  # xAI
  "grok-4.20",
  # Meta
  "llama-3.1-8b", "llama-3.1-70b",
  # Qwen
  "qwen3-8b", "qwen3-8b-reasoning", "qwen3-32b", "qwen3-32b-reasoning"
)

MODEL_FAMILY <- c(
  "gpt-4o" = "OpenAI", "gpt-5" = "OpenAI", "gpt-5.3" = "OpenAI", "gpt-5.3-reasoning" = "OpenAI",
  "claude-sonnet-3.7" = "Anthropic", "claude-sonnet-4.6" = "Anthropic",
  "claude-opus-4.6" = "Anthropic", "claude-opus-4.6-reasoning" = "Anthropic",
  "gemini-2.5-pro" = "Google", "grok-4.20" = "xAI",
  "llama-3.1-8b" = "Meta", "llama-3.1-70b" = "Meta",
  "qwen3-8b" = "Qwen", "qwen3-8b-reasoning" = "Qwen",
  "qwen3-32b" = "Qwen", "qwen3-32b-reasoning" = "Qwen"
)

REASONING_PAIRS <- tribble(
  ~base,                  ~reasoning,
  "claude-opus-4.6",      "claude-opus-4.6-reasoning",
  "gpt-5.3",              "gpt-5.3-reasoning",
  "qwen3-8b",             "qwen3-8b-reasoning",
  "qwen3-32b",            "qwen3-32b-reasoning"
)

# ---- Load -------------------------------------------------------------------
read_judged <- function(path) {
  model <- str_replace(basename(path), "^responses_(.*)\\.judged\\.csv$", "\\1")
  suppressWarnings(readr::read_csv(
    path, show_col_types = FALSE,
    col_types = cols(.default = col_character())
  )) |>
    mutate(model = model)
}

raw <- list.files(ANNOTATION_DIR, pattern = "^responses_.*\\.judged\\.csv$",
                  full.names = TRUE) |>
  map_dfr(read_judged)

# Coerce binary L2 columns to integer where present, leaving NA elsewhere.
to_int_binary <- function(x) {
  x <- na_if(trimws(x), "")
  suppressWarnings(as.integer(x))
}
# judge_skipped only appears as a column when at least one row in that file
# was skipped; ensure it always exists for downstream code.
if (!"judge_skipped" %in% names(raw)) raw$judge_skipped <- NA_character_
if (!"skip_reason"  %in% names(raw)) raw$skip_reason  <- NA_character_

df <- raw |>
  mutate(across(all_of(L2_FEATURES), to_int_binary),
         parse_error   = if_else(parse_error   %in% c("True","TRUE","1"), TRUE, FALSE, missing = FALSE),
         judge_skipped = if_else(judge_skipped %in% c("True","TRUE","1"), TRUE, FALSE, missing = FALSE)) |>
  mutate(model = factor(model, levels = MODEL_ORDER),
         family = factor(MODEL_FAMILY[as.character(model)],
                         levels = c("OpenAI","Anthropic","Google","xAI","Meta","Qwen")))

# Defensive: drop rows that don't match the current prompt SHA256, EXCEPT skips
# (which never reach the judge so don't carry the hash).
df <- df |>
  filter(is.na(judge_prompt_sha256) |
           judge_prompt_sha256 == "" |
           judge_prompt_sha256 == PROMPT_SHA256)

# Convenience splits.
df_clean <- df |> filter(!parse_error, !is.na(layer0))   # successful judge rows
df_nc    <- df_clean |> filter(layer0 == "Non-compliance")
```

# Overview

This document characterises the refusal behaviour of `r length(unique(df$model))`
LLMs across `r nrow(df)` prompt-response pairs sampled from LMSYS/WildChat and
SorryBench, judged by GPT-5.5 (non-reasoning) using the codebook-aligned prompt
(SHA256 prefix `r substr(PROMPT_SHA256, 1, 12)`...).

It is descriptive scaffolding for the paper — not the final analysis. Outputs
are organised as:

1. **Coverage** — how many records per model reached the judge.
2. **Layer 0 (Action)** — Full / Partial / Non-compliance distribution.
3. **Layer 1 (Rationale)** — Bare / Capacity / Policy / Ethics, within NC.
4. **Layer 2 (Form features)** — prevalence by feature, within NC.
5. **Harm category breakdowns** — Layer 0/1/2 by Llama Guard category.
6. **Cross-cutting analyses** — reasoning vs non-reasoning pairs, model families.
7. **Flavour** — sampled responses per model and category, for qualitative read.

---

# 1. Coverage

```{r coverage}
cov <- df |>
  group_by(model) |>
  summarise(
    n_total     = n(),
    n_judged    = sum(!parse_error & !judge_skipped & !is.na(layer0)),
    n_skipped   = sum(judge_skipped),
    n_parse_err = sum(parse_error),
    .groups = "drop"
  ) |>
  mutate(pct_judged = percent(n_judged / n_total, accuracy = 0.1))

cov |>
  kable(caption = "Per-model coverage. Skips here include upstream provider blocks (Gemini/Grok), Claude 4.6 refusal-stop signals, and transient inference failures.") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# 2. Layer 0: Action distribution

## 2.1 Per-model distribution

```{r l0-by-model, fig.height = 7}
l0_levels <- c("Full compliance", "Partial compliance", "Non-compliance")
l0_pal    <- c("Full compliance" = "#1b9e77",
               "Partial compliance" = "#d95f02",
               "Non-compliance" = "#7570b3")

l0_tbl <- df_clean |>
  count(model, layer0) |>
  group_by(model) |>
  mutate(prop = n / sum(n),
         layer0 = factor(layer0, levels = l0_levels)) |>
  ungroup()

ggplot(l0_tbl, aes(model, prop, fill = layer0)) +
  geom_col(width = 0.8) +
  scale_y_continuous(labels = percent_format(1)) +
  scale_fill_manual(values = l0_pal) +
  coord_flip() +
  labs(x = NULL, y = NULL, fill = "Layer 0",
       title = "Action distribution by model",
       subtitle = "Among successfully judged responses") +
  theme_minimal(base_size = 12)
```

## 2.2 Refusal rate (NC) ranking

```{r refusal-rate}
df_clean |>
  group_by(model) |>
  summarise(n = n(),
            refusal_rate = mean(layer0 == "Non-compliance"),
            partial_rate = mean(layer0 == "Partial compliance"),
            comply_rate  = mean(layer0 == "Full compliance"),
            .groups = "drop") |>
  arrange(desc(refusal_rate)) |>
  mutate(across(ends_with("_rate"), ~percent(.x, accuracy = 0.1))) |>
  kable(caption = "Models ranked by refusal rate.") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# 3. Layer 1: Refusal basis (among Non-compliance)

```{r l1-by-model, fig.height = 7}
l1_levels <- c("Bare", "Capacity-based", "Policy-based", "Ethics-based")
l1_pal    <- c("Bare" = "#bdbdbd",
               "Capacity-based" = "#fdae61",
               "Policy-based" = "#abd9e9",
               "Ethics-based" = "#2c7bb6")

l1_tbl <- df_nc |>
  filter(layer1 %in% l1_levels) |>
  count(model, layer1) |>
  group_by(model) |>
  mutate(prop = n / sum(n),
         layer1 = factor(layer1, levels = l1_levels)) |>
  ungroup()

ggplot(l1_tbl, aes(model, prop, fill = layer1)) +
  geom_col(width = 0.8) +
  scale_y_continuous(labels = percent_format(1)) +
  scale_fill_manual(values = l1_pal) +
  coord_flip() +
  labs(x = NULL, y = NULL, fill = "Layer 1",
       title = "Refusal basis among Non-compliance",
       subtitle = "Proportional within each model's NC responses") +
  theme_minimal(base_size = 12)
```

```{r l1-table}
df_nc |>
  filter(layer1 %in% l1_levels) |>
  count(model, layer1) |>
  group_by(model) |>
  mutate(pct = percent(n / sum(n), accuracy = 0.1)) |>
  select(model, layer1, n, pct) |>
  pivot_wider(names_from = layer1, values_from = c(n, pct), values_fill = list(n = 0, pct = "0%")) |>
  kable(caption = "Layer 1 distribution by model.") |>
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 11)
```

---

# 4. Layer 2: Form features (among Non-compliance)

## 4.1 Feature prevalence heatmap

```{r l2-heatmap, fig.height = 7}
l2_tbl <- df_nc |>
  select(model, all_of(L2_FEATURES)) |>
  pivot_longer(-model, names_to = "feature", values_to = "value") |>
  filter(!is.na(value)) |>
  group_by(model, feature) |>
  summarise(rate = mean(value), n = n(), .groups = "drop") |>
  mutate(feature = factor(feature, levels = L2_FEATURES))

ggplot(l2_tbl, aes(feature, model, fill = rate)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  geom_text(aes(label = percent(rate, accuracy = 1)), size = 3) +
  scale_fill_gradient(low = "#f7f7f7", high = "#08519c",
                      labels = percent_format(1), name = "Rate") +
  scale_x_discrete(position = "top") +
  labs(x = NULL, y = NULL,
       title = "L2 feature rates by model (within Non-compliance)") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 0))
```

## 4.2 Feature-by-feature ranking

```{r l2-ranking}
l2_tbl |>
  group_by(feature) |>
  summarise(
    overall = percent(weighted.mean(rate, n), accuracy = 0.1),
    top_model = model[which.max(rate)],
    top_rate  = percent(max(rate), accuracy = 0.1),
    bot_model = model[which.min(rate)],
    bot_rate  = percent(min(rate), accuracy = 0.1),
    .groups = "drop"
  ) |>
  kable(caption = "Where each L2 feature peaks and bottoms out.") |>
  kable_styling(bootstrap_options = c("striped","hover"), font_size = 11)
```

---

# 5. By harm category

## 5.1 L0 by category × model

```{r l0-by-cat, fig.height = 9}
df_clean |>
  count(primary_llama_guard_category_name, model, layer0) |>
  group_by(primary_llama_guard_category_name, model) |>
  mutate(prop = n / sum(n),
         layer0 = factor(layer0, levels = l0_levels)) |>
  ungroup() |>
  ggplot(aes(model, prop, fill = layer0)) +
    geom_col(width = 0.8) +
    coord_flip() +
    facet_wrap(~ primary_llama_guard_category_name, ncol = 3) +
    scale_y_continuous(labels = percent_format(1), breaks = c(0, 0.5, 1)) +
    scale_fill_manual(values = l0_pal) +
    labs(x = NULL, y = NULL, fill = "Layer 0",
         title = "Action distribution: model × harm category") +
    theme_minimal(base_size = 10) +
    theme(legend.position = "bottom",
          strip.text = element_text(face = "bold", size = 9))
```

## 5.2 Refusal rate by category

```{r refusal-rate-by-cat}
df_clean |>
  group_by(primary_llama_guard_category_name) |>
  summarise(n = n(),
            mean_refusal = mean(layer0 == "Non-compliance"),
            sd_across_models = sd(tapply(layer0 == "Non-compliance", model, mean), na.rm = TRUE),
            .groups = "drop") |>
  arrange(desc(mean_refusal)) |>
  mutate(across(c(mean_refusal, sd_across_models), ~percent(.x, accuracy = 0.1))) |>
  kable(caption = "Harm categories ranked by mean refusal rate across models. SD indicates between-model disagreement.") |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

## 5.3 L1 rationale shifts across categories

```{r l1-by-cat}
df_nc |>
  filter(layer1 %in% l1_levels) |>
  count(primary_llama_guard_category_name, layer1) |>
  group_by(primary_llama_guard_category_name) |>
  mutate(prop = n / sum(n),
         layer1 = factor(layer1, levels = l1_levels)) |>
  ungroup() |>
  ggplot(aes(primary_llama_guard_category_name, prop, fill = layer1)) +
    geom_col(width = 0.8) +
    scale_y_continuous(labels = percent_format(1)) +
    scale_fill_manual(values = l1_pal) +
    coord_flip() +
    labs(x = NULL, y = NULL, fill = "Layer 1",
         title = "L1 rationale by harm category (pooled across models)") +
    theme_minimal(base_size = 11)
```

## 5.4 L2 features by category (selected)

For preliminary exploration: which categories elicit hedging, solidarity, and
normative suggestion? These are the features Ruoxuan flagged as most
diagnostic of human-like vs safety-classifier-like refusals.

```{r l2-by-cat-selected, fig.height = 7}
selected_features <- c("apology", "solidarity", "negative_stance",
                       "normative_suggestion", "alternative_offer", "executed_alternative")

df_nc |>
  select(primary_llama_guard_category_name, all_of(selected_features)) |>
  pivot_longer(-primary_llama_guard_category_name, names_to = "feature", values_to = "value") |>
  filter(!is.na(value)) |>
  group_by(primary_llama_guard_category_name, feature) |>
  summarise(rate = mean(value), .groups = "drop") |>
  mutate(feature = factor(feature, levels = selected_features)) |>
  ggplot(aes(feature, primary_llama_guard_category_name, fill = rate)) +
    geom_tile(colour = "white", linewidth = 0.3) +
    geom_text(aes(label = percent(rate, accuracy = 1)), size = 2.8) +
    scale_fill_gradient(low = "#f7f7f7", high = "#08519c",
                        labels = percent_format(1)) +
    labs(x = NULL, y = NULL, fill = "Rate",
         title = "Selected L2 features × harm category") +
    theme_minimal(base_size = 10) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

---

# 6. Cross-cutting analyses

## 6.1 Reasoning vs non-reasoning

For models with both a base and a reasoning variant: does turning on the
thinking trace change the refusal profile?

```{r reasoning-pairs}
pair_tbl <- df_clean |>
  filter(model %in% c(REASONING_PAIRS$base, REASONING_PAIRS$reasoning)) |>
  mutate(variant = if_else(model %in% REASONING_PAIRS$reasoning, "reasoning", "base"),
         base_model = case_when(
           model %in% REASONING_PAIRS$base ~ as.character(model),
           model %in% REASONING_PAIRS$reasoning ~ str_replace(as.character(model), "-reasoning$", ""),
           TRUE ~ NA_character_
         )) |>
  group_by(base_model, variant) |>
  summarise(
    refusal_rate = mean(layer0 == "Non-compliance"),
    partial_rate = mean(layer0 == "Partial compliance"),
    .groups = "drop"
  ) |>
  pivot_longer(ends_with("_rate"), names_to = "metric", values_to = "rate") |>
  pivot_wider(names_from = variant, values_from = rate) |>
  mutate(delta = reasoning - base)

pair_tbl |>
  mutate(across(c(base, reasoning, delta), ~percent(.x, accuracy = 0.1))) |>
  kable(caption = "Reasoning vs base: Δ in refusal/partial rate. Positive Δ means reasoning variant refuses more.") |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

```{r reasoning-l2-deltas, fig.height = 5}
df_clean |>
  filter(layer0 == "Non-compliance",
         model %in% c(REASONING_PAIRS$base, REASONING_PAIRS$reasoning)) |>
  mutate(variant = if_else(model %in% REASONING_PAIRS$reasoning, "reasoning", "base"),
         base_model = if_else(model %in% REASONING_PAIRS$reasoning,
                              str_replace(as.character(model), "-reasoning$", ""),
                              as.character(model))) |>
  select(base_model, variant, all_of(L2_FEATURES)) |>
  pivot_longer(-c(base_model, variant), names_to = "feature", values_to = "value") |>
  filter(!is.na(value)) |>
  group_by(base_model, variant, feature) |>
  summarise(rate = mean(value), .groups = "drop") |>
  pivot_wider(names_from = variant, values_from = rate) |>
  mutate(delta = reasoning - base,
         feature = factor(feature, levels = L2_FEATURES)) |>
  ggplot(aes(feature, delta, fill = base_model)) +
    geom_col(position = position_dodge(width = 0.8), width = 0.8) +
    geom_hline(yintercept = 0, colour = "grey40") +
    scale_y_continuous(labels = percent_format(1)) +
    labs(x = NULL, y = "Δ (reasoning − base)", fill = "Model pair",
         title = "L2 feature shifts when reasoning is enabled") +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "bottom")
```

## 6.2 Model family comparison

```{r family-l0}
df_clean |>
  group_by(family) |>
  summarise(
    n_models = n_distinct(model),
    n_records = n(),
    refusal_rate = mean(layer0 == "Non-compliance"),
    partial_rate = mean(layer0 == "Partial compliance"),
    full_rate    = mean(layer0 == "Full compliance"),
    .groups = "drop"
  ) |>
  mutate(across(ends_with("_rate"), ~percent(.x, accuracy = 0.1))) |>
  kable(caption = "Layer 0 distribution by provider family (pooled).") |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

```{r family-l2, fig.height = 6}
df_nc |>
  select(family, all_of(L2_FEATURES)) |>
  pivot_longer(-family, names_to = "feature", values_to = "value") |>
  filter(!is.na(value)) |>
  group_by(family, feature) |>
  summarise(rate = mean(value), .groups = "drop") |>
  mutate(feature = factor(feature, levels = L2_FEATURES)) |>
  ggplot(aes(feature, family, fill = rate)) +
    geom_tile(colour = "white", linewidth = 0.4) +
    geom_text(aes(label = percent(rate, accuracy = 1)), size = 3) +
    scale_fill_gradient(low = "#f7f7f7", high = "#08519c", labels = percent_format(1)) +
    labs(x = NULL, y = NULL, fill = "Rate",
         title = "L2 feature rates by provider family") +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## 6.3 Selected hypothesis-shaped views

> *Pre-registered framing direction (paper draft): modern safety training appears
> to shift LLM refusals toward firmer boundaries and away from human-like
> face-saving moves. The plots below set up that contrast — they are descriptive
> only and intentionally avoid significance testing at this stage.*

### Apology decay across OpenAI and Anthropic releases

```{r apology-decay, fig.height = 4}
df_nc |>
  filter(family %in% c("OpenAI", "Anthropic"),
         !is.na(apology)) |>
  group_by(family, model) |>
  summarise(apology_rate = mean(apology), n = n(), .groups = "drop") |>
  ggplot(aes(model, apology_rate, fill = family)) +
    geom_col() +
    geom_text(aes(label = percent(apology_rate, accuracy = 1)), vjust = -0.4, size = 3) +
    scale_y_continuous(labels = percent_format(1), expand = expansion(mult = c(0, 0.1))) +
    labs(x = NULL, y = "apology rate",
         title = "Apology in non-compliance: OpenAI vs Anthropic",
         subtitle = "Earlier models lean on apology; newer ones do not") +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "bottom")
```

### Negative stance is high almost everywhere

```{r negstance-prevalence, fig.height = 4}
df_nc |>
  filter(!is.na(negative_stance)) |>
  group_by(family, model) |>
  summarise(rate = mean(negative_stance), .groups = "drop") |>
  ggplot(aes(model, rate, fill = family)) +
    geom_col() +
    geom_text(aes(label = percent(rate, accuracy = 1)), vjust = -0.4, size = 3) +
    scale_y_continuous(labels = percent_format(1), expand = expansion(mult = c(0, 0.1))) +
    labs(x = NULL, y = "negative_stance rate",
         title = "Negative stance prevalence by model") +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "bottom")
```

### Affiliative features: positive alignment & solidarity

```{r affiliative, fig.height = 5}
df_nc |>
  select(family, model, positive_alignment, solidarity) |>
  pivot_longer(c(positive_alignment, solidarity), names_to = "feature", values_to = "value") |>
  filter(!is.na(value)) |>
  group_by(family, model, feature) |>
  summarise(rate = mean(value), .groups = "drop") |>
  ggplot(aes(model, rate, fill = feature)) +
    geom_col(position = position_dodge(width = 0.8), width = 0.8) +
    scale_y_continuous(labels = percent_format(1)) +
    labs(x = NULL, y = NULL, fill = NULL,
         title = "Affiliative refusal features by model") +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "bottom")
```

---

# 7. Flavour: sampled responses per model

This section samples a small number of responses per (model × Layer 0 outcome)
so the reader can ground numerical findings in concrete language. Responses are
**not cherry-picked**: they are drawn at random with a fixed seed inside this
chunk. Long responses are truncated to keep the document readable.

> **Note.** This section will surface the raw user queries and model responses,
> which include harmful prompts from SorryBench/LMSYS. View with awareness; the
> code generates the sample without anyone needing to read the corpus first.

```{r flavour-config}
set.seed(20260519)
MAX_QUERY_CHARS    <- 400
MAX_RESPONSE_CHARS <- 900
SAMPLES_PER_CELL   <- 1   # per (model × L0). Bump to 2-3 once layout is reviewed.
```

## 7.1 One example per (model × L0)

```{r flavour-by-model}
truncate_text <- function(x, n) {
  x <- as.character(x)
  ifelse(nchar(x) > n, paste0(substr(x, 1, n - 3), "..."), x)
}

flavour <- df_clean |>
  filter(!is.na(query), !is.na(response), nchar(response) > 0) |>
  group_by(model, layer0) |>
  slice_sample(n = SAMPLES_PER_CELL) |>
  ungroup() |>
  arrange(model, factor(layer0, levels = l0_levels)) |>
  mutate(
    query    = truncate_text(query,    MAX_QUERY_CHARS),
    response = truncate_text(response, MAX_RESPONSE_CHARS),
    L2 = pmap_chr(
      select(., all_of(L2_FEATURES)),
      function(...) {
        vals <- list(...)
        names(vals) <- L2_FEATURES
        on <- names(vals)[vapply(vals, function(v) isTRUE(as.integer(v) == 1L), logical(1))]
        if (length(on) == 0) "—" else paste(on, collapse = ", ")
      }
    )
  ) |>
  select(model, layer0, layer1, L2, category = primary_llama_guard_category_name, query, response)

flavour |>
  kable(caption = "One sampled response per (model × Layer 0 outcome).") |>
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 10) |>
  column_spec(6, width = "20em") |>
  column_spec(7, width = "30em")
```

## 7.2 One example per (harm category × representative model)

To get a flavour across the request spectrum: for each harm category, sample
one response from `gpt-4o`, `claude-sonnet-4.6`, `gemini-2.5-pro`, and
`llama-3.1-70b` (a spread of provider safety stances).

```{r flavour-by-category}
focal_models <- c("gpt-4o", "claude-sonnet-4.6", "gemini-2.5-pro", "llama-3.1-70b")

df_clean |>
  filter(model %in% focal_models,
         !is.na(query), !is.na(response), nchar(response) > 0) |>
  group_by(primary_llama_guard_category_name, model) |>
  slice_sample(n = 1) |>
  ungroup() |>
  arrange(primary_llama_guard_category_name, factor(model, levels = focal_models)) |>
  mutate(
    query    = truncate_text(query, MAX_QUERY_CHARS),
    response = truncate_text(response, MAX_RESPONSE_CHARS)
  ) |>
  select(category = primary_llama_guard_category_name, model, layer0, layer1,
         query, response) |>
  kable(caption = "One sampled response per harm category × focal model.") |>
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 10) |>
  column_spec(5, width = "20em") |>
  column_spec(6, width = "30em")
```

---

# 8. Statistical hypothesis tests (scaffolding)

> EMNLP papers usually carry their argument with descriptive plots and a small
> number of well-chosen tests rather than a full stats appendix. The blocks
> below give a working starting point — chi-square omnibus tests, McNemar for
> the paired reasoning comparisons, a mixed-effects model that respects the
> nested structure (prompts × models), and a composite "human-like mitigation"
> score. Adapt freely; report what survives multiple-comparison correction.

```{r stats-setup}
# Optional packages; the chunk degrades gracefully if missing.
has_lme4   <- requireNamespace("lme4",   quietly = TRUE)
has_broomm <- requireNamespace("broom.mixed", quietly = TRUE)
if (has_lme4)   suppressMessages(library(lme4))
if (has_broomm) suppressMessages(library(broom.mixed))

# Prompt identifier shared across models (same 200 prompts hit every model).
df_clean <- df_clean |>
  mutate(prompt_id = coalesce(final_sample_id, candidate_id))
df_nc <- df_nc |>
  mutate(prompt_id = coalesce(final_sample_id, candidate_id))
```

## 8.1 Omnibus: do models differ?

```{r omnibus-model-l0}
tab_model_l0 <- df_clean |>
  count(model, layer0) |>
  pivot_wider(names_from = layer0, values_from = n, values_fill = 0) |>
  column_to_rownames("model") |>
  as.matrix()

chi_model <- chisq.test(tab_model_l0)
cramers_v <- sqrt(chi_model$statistic /
                  (sum(tab_model_l0) * (min(dim(tab_model_l0)) - 1)))

tibble(
  test = "χ²(L0 × model)",
  statistic = round(chi_model$statistic, 2),
  df = chi_model$parameter,
  p_value = format.pval(chi_model$p.value, digits = 3),
  cramers_v = round(as.numeric(cramers_v), 3)
) |>
  kable(caption = "Models differ in Layer 0 outcome (omnibus). Cramér's V quantifies the effect size; >0.1 small, >0.3 medium, >0.5 large.") |>
  kable_styling(full_width = FALSE)
```

## 8.2 Omnibus: do harm categories elicit different actions?

```{r omnibus-cat-l0}
tab_cat_l0 <- df_clean |>
  count(primary_llama_guard_category_name, layer0) |>
  pivot_wider(names_from = layer0, values_from = n, values_fill = 0) |>
  column_to_rownames("primary_llama_guard_category_name") |>
  as.matrix()

chi_cat <- chisq.test(tab_cat_l0)
v_cat   <- sqrt(chi_cat$statistic /
                (sum(tab_cat_l0) * (min(dim(tab_cat_l0)) - 1)))

tibble(
  test = "χ²(L0 × harm category)",
  statistic = round(chi_cat$statistic, 2),
  df = chi_cat$parameter,
  p_value = format.pval(chi_cat$p.value, digits = 3),
  cramers_v = round(as.numeric(v_cat), 3)
) |>
  kable(caption = "Harm categories differ in elicited Layer 0 outcome (pooled across models).") |>
  kable_styling(full_width = FALSE)
```

## 8.3 Reasoning vs base: McNemar paired tests

The same 200 prompts are sent to both the base and reasoning variant of a
model, so the right test for "does turning on reasoning change refusal
behaviour?" is paired. McNemar's test on the discordant cells of the 2×2
table is the standard choice for a binary outcome.

```{r mcnemar-l0}
mcnemar_one <- function(base_model, reasoning_model, outcome_col, outcome_value) {
  pair <- df_clean |>
    filter(model %in% c(base_model, reasoning_model)) |>
    mutate(variant = if_else(model == reasoning_model, "reasoning", "base"),
           hit = .data[[outcome_col]] == outcome_value) |>
    select(prompt_id, variant, hit) |>
    pivot_wider(names_from = variant, values_from = hit, values_fn = first) |>
    drop_na(base, reasoning)

  tab <- table(base = pair$base, reasoning = pair$reasoning)
  if (any(dim(tab) < 2)) return(tibble(test = NA, statistic = NA, p_value = NA, n = nrow(pair)))
  m <- mcnemar.test(tab)
  tibble(statistic = unname(m$statistic), p_value = m$p.value, n = nrow(pair))
}

l0_pairs <- REASONING_PAIRS |>
  rowwise() |>
  mutate(res = list(mcnemar_one(base, reasoning, "layer0", "Non-compliance"))) |>
  unnest(res) |>
  ungroup() |>
  mutate(p_bh = p.adjust(p_value, "BH"))

l0_pairs |>
  mutate(p_value = format.pval(p_value, digits = 3),
         p_bh    = format.pval(p_bh,    digits = 3),
         statistic = round(statistic, 2)) |>
  kable(caption = "McNemar test on `refused (NC) vs not` for each base/reasoning pair.") |>
  kable_styling(full_width = FALSE)
```

```{r mcnemar-l2}
# Same logic, applied to every L2 feature × pair.
mcnemar_feature <- function(base_model, reasoning_model, feature) {
  pair <- df_clean |>
    filter(model %in% c(base_model, reasoning_model),
           !is.na(.data[[feature]])) |>
    mutate(variant = if_else(model == reasoning_model, "reasoning", "base"),
           hit = .data[[feature]] == 1) |>
    select(prompt_id, variant, hit) |>
    pivot_wider(names_from = variant, values_from = hit, values_fn = first) |>
    drop_na(base, reasoning)

  tab <- table(base = pair$base, reasoning = pair$reasoning)
  if (any(dim(tab) < 2) || nrow(pair) < 10)
    return(tibble(statistic = NA_real_, p_value = NA_real_, n = nrow(pair)))
  m <- mcnemar.test(tab)
  tibble(statistic = unname(m$statistic), p_value = m$p.value, n = nrow(pair))
}

l2_pairs <- REASONING_PAIRS |>
  crossing(feature = L2_FEATURES) |>
  rowwise() |>
  mutate(res = list(mcnemar_feature(base, reasoning, feature))) |>
  unnest(res) |>
  ungroup() |>
  group_by(base) |>
  mutate(p_bh = p.adjust(p_value, "BH")) |>
  ungroup()

l2_pairs |>
  filter(!is.na(p_value)) |>
  mutate(p_value = format.pval(p_value, digits = 3),
         p_bh    = format.pval(p_bh,    digits = 3),
         statistic = round(statistic, 2)) |>
  arrange(base, factor(feature, levels = L2_FEATURES)) |>
  kable(caption = "L2 feature shifts under reasoning, paired McNemar with BH-corrected p-values within each model pair.") |>
  kable_styling(full_width = FALSE, font_size = 10)
```

## 8.4 Mixed-effects model: model effect on refusal, controlling for prompt

A chi-square treats every (model × prompt) cell as independent, which they
aren't — the same 200 prompts hit all 16 models. A logistic mixed model with
a random intercept per prompt is the cleanest "model X refuses more than
model Y, controlling for prompt difficulty" estimator.

```{r glmm-refusal, eval = has_lme4}
fit_data <- df_clean |>
  mutate(refused = as.integer(layer0 == "Non-compliance"),
         model   = factor(model, levels = MODEL_ORDER),
         prompt_id = factor(prompt_id)) |>
  filter(!is.na(refused), !is.na(prompt_id))

# Reference level: gpt-4o (oldest and most permissive OpenAI in the set).
fit_data$model <- relevel(fit_data$model, ref = "gpt-4o")

m_refuse <- glmer(refused ~ model + (1 | prompt_id),
                  data = fit_data,
                  family = binomial,
                  control = glmerControl(optimizer = "bobyqa"))

if (has_broomm) {
  broom.mixed::tidy(m_refuse, effects = "fixed", conf.int = TRUE, exponentiate = TRUE) |>
    filter(term != "(Intercept)") |>
    mutate(term = str_replace(term, "^model", ""),
           across(c(estimate, conf.low, conf.high), ~round(.x, 2)),
           p.value = format.pval(p.value, digits = 3)) |>
    select(model = term, OR = estimate, ci_lo = conf.low, ci_hi = conf.high, p = p.value) |>
    kable(caption = "Odds ratios for refusal vs gpt-4o, from glmer(refused ~ model + (1|prompt)). OR > 1 means the model refuses more than gpt-4o on a given prompt.") |>
    kable_styling(full_width = FALSE, font_size = 11)
} else {
  print(summary(m_refuse))
}
```

```{r glmm-refusal-fallback, eval = !has_lme4, echo = FALSE, results = 'asis'}
cat("> `lme4` not installed; mixed-effects table skipped. Install with `install.packages('lme4')` to enable.\n")
```

## 8.5 Composite "human-like mitigation" score

Pre-registered framing: modern safety training shifts refusals away from
human-like face-saving moves. Operationalise a per-response score as the
count of affiliative + mitigating features present (range 0–4):

```{r composite-score}
mitigation_features <- c("apology", "hedge", "positive_alignment", "solidarity")

df_nc <- df_nc |>
  mutate(mitigation_score = rowSums(across(all_of(mitigation_features)), na.rm = TRUE))

df_nc |>
  group_by(family, model) |>
  summarise(n = n(),
            mean_score = mean(mitigation_score, na.rm = TRUE),
            sd_score   = sd(mitigation_score, na.rm = TRUE),
            .groups = "drop") |>
  arrange(desc(mean_score)) |>
  mutate(across(c(mean_score, sd_score), ~round(.x, 2))) |>
  kable(caption = "Mean count of affiliative/mitigating features per NC response (max 4).") |>
  kable_styling(full_width = FALSE, font_size = 11)
```

```{r composite-by-family}
# One-way comparison of mitigation_score across families.
kw_family <- kruskal.test(mitigation_score ~ family, data = df_nc)
tibble(
  test = "Kruskal–Wallis: mitigation_score ~ family",
  statistic = round(kw_family$statistic, 2),
  df = kw_family$parameter,
  p_value = format.pval(kw_family$p.value, digits = 3)
) |>
  kable() |>
  kable_styling(full_width = FALSE)
```

```{r composite-mixed, eval = has_lme4}
# Mixed model on the composite: family effect with prompt random intercept.
fit_score <- df_nc |>
  mutate(prompt_id = factor(prompt_id),
         family    = factor(family))

m_score <- lmer(mitigation_score ~ family + (1 | prompt_id), data = fit_score)
if (has_broomm) {
  broom.mixed::tidy(m_score, effects = "fixed", conf.int = TRUE) |>
    mutate(term = str_replace(term, "^family", ""),
           across(c(estimate, conf.low, conf.high), ~round(.x, 2))) |>
    kable(caption = "Mixed model: mitigation_score ~ family + (1|prompt). Reference family = OpenAI.") |>
    kable_styling(full_width = FALSE)
} else {
  print(summary(m_score))
}
```

## 8.6 Caveats to flag in any reported test

- **Independence.** The same 200 prompts hit every model; pooled chi-squares
  understate uncertainty. Where models are the unit of analysis, prefer the
  mixed-effects estimates in 8.4 / 8.5.
- **Judge as a fixed effect.** All results are conditional on GPT-5.5's
  judgements at this prompt. Sensitivity analysis on a different judge model
  is owed before any test result becomes a paper claim.
- **Multiple comparisons.** L2 has 13 binary features × 4 reasoning pairs =
  52 paired tests; the section already applies BH within each pair. For
  cross-model L2 comparisons (one feature, 16 models, 120 pairwise tests),
  prefer omnibus + a few principled contrasts over exhaustive pairwise.
- **Provider safety blocks.** Gemini's 19 Google-blocked prompts and Grok's
  19 xAI-blocked prompts are excluded from `df_clean` here. If you want to
  count them as system-level NC, recompute with those rows added back —
  they will shift refusal rates and L1 distributions for those two models.

---

# Appendix: Data sources

- Judged data: `data/annotations/responses_<model>.judged.csv`, 16 model files.
- Judge: `openai/gpt-5.5` non-reasoning, prompt SHA256 prefix
  `r substr(PROMPT_SHA256, 1, 12)`...
- Codebook: `taxonomy_guideline (11).pdf`.
- Prompts: 200 sampled from LMSYS/WildChat (`source = "lmsys"`) and
  SorryBench (`source = "sorrybench"`), balanced across Llama Guard harm
  categories.
- Skips (provider-blocked / refusal-stop / transient) and the run-1 OR-credit
  parse_errors are documented in `logs/` and excluded from the analyses above.
