---
title: "LLM Refusal Strategy — Descriptive Analysis"
author: "Polite Refusals team"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    highlight: kate
    theme: flatly
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 6, dpi = 110
)
library(tidyverse)
library(jsonlite)
library(scales)
library(knitr)
library(kableExtra)

# ---- Config -----------------------------------------------------------------
ANNOTATION_DIR <- file.path("..", "data", "annotations")
# Current judge prompt SHA256 (from BASE_SYSTEM_PROMPT in scripts/utils/judge_config.py).
# Records whose judge_prompt_sha256 != this are dropped, defending against
# accidental mixing of runs after future prompt revisions.
PROMPT_SHA256 <- "238bb923037bb4126bef041f202fd6547e328aa62f4f61f06b8f63d784891fb1"

L2_FEATURES <- c(
  "explicit_nc", "implicit_nc", "apology", "hedge", "explanatory_preface",
  "positive_alignment", "solidarity", "negative_stance",
  "executed_alternative", "alternative_offer", "normative_suggestion",
  "statement_of_principle", "role_based"
)

# Model display order: group by family, oldest → newest within family so the
# legacy/older models (added 2026-05-20 for temporal coverage) sit at the head
# of each family group.
MODEL_ORDER <- c(
  # OpenAI: legacy → modern
  #"gpt-4-turbo", 
  "gpt-4o", "gpt-5", "gpt-5.3", "gpt-5.3-reasoning",
  # Anthropic: legacy → modern
  "claude-sonnet-3.7", "claude-sonnet-4.6", "claude-opus-4.6", "claude-opus-4.6-reasoning",
  # Google
  "gemini-2.5-pro",
  # xAI
  "grok-4.20",
  # Meta: legacy → modern
  #"llama-3-70b-instruct", 
  "llama-3.1-8b", "llama-3.1-70b",
  # Mistral
  #"mistral-7b-instruct-v0.1",
  # Qwen
  "qwen3-8b", "qwen3-8b-reasoning", "qwen3-32b", "qwen3-32b-reasoning"
)

MODEL_FAMILY <- c(
  # "gpt-4-turbo" = "OpenAI",
  "gpt-4o" = "OpenAI", "gpt-5" = "OpenAI", "gpt-5.3" = "OpenAI", "gpt-5.3-reasoning" = "OpenAI",
  "claude-sonnet-3.7" = "Anthropic", "claude-sonnet-4.6" = "Anthropic",
  "claude-opus-4.6" = "Anthropic", "claude-opus-4.6-reasoning" = "Anthropic",
  "gemini-2.5-pro" = "Google", "grok-4.20" = "xAI",
  # "llama-3-70b-instruct" = "Meta",
  "llama-3.1-8b" = "Meta", "llama-3.1-70b" = "Meta",
  # "mistral-7b-instruct-v0.1" = "Mistral",
  "qwen3-8b" = "Qwen", "qwen3-8b-reasoning" = "Qwen",
  "qwen3-32b" = "Qwen", "qwen3-32b-reasoning" = "Qwen"
)

REASONING_PAIRS <- tribble(
  ~base,                  ~reasoning,
  "claude-opus-4.6",      "claude-opus-4.6-reasoning",
  "gpt-5.3",              "gpt-5.3-reasoning",
  "qwen3-8b",             "qwen3-8b-reasoning",
  "qwen3-32b",            "qwen3-32b-reasoning"
)

# ---- Load -------------------------------------------------------------------
# Older models (gpt-4-turbo, llama-3-70b-instruct, mistral-7b-instruct-v0.1)
# were collected for temporal-coverage analysis but are excluded from the main
# corpus: they sit awkwardly in the corpus (different generation, weaker
# instruction-following, distinct release-era), and the within-family
# old-vs-new contrasts were better served by the gpt-4o/gpt-5 and
# sonnet-3.7/sonnet-4.6 pairs already in scope. Data files remain on disk.
EXCLUDED_MODELS <- c("gpt-4-turbo", "llama-3-70b-instruct", "mistral-7b-instruct-v0.1")

read_judged <- function(path) {
  model <- str_replace(basename(path), "^responses_(.*)\\.judged\\.csv$", "\\1")
  suppressWarnings(readr::read_csv(
    path, show_col_types = FALSE,
    col_types = cols(.default = col_character())
  )) |>
    mutate(model = model)
}

raw <- list.files(ANNOTATION_DIR, pattern = "^responses_.*\\.judged\\.csv$",
                  full.names = TRUE) |>
  map_dfr(read_judged) |>
  filter(!model %in% EXCLUDED_MODELS)

# Coerce binary L2 columns to integer where present, leaving NA elsewhere.
to_int_binary <- function(x) {
  x <- na_if(trimws(x), "")
  suppressWarnings(as.integer(x))
}
# judge_skipped only appears as a column when at least one row in that file
# was skipped; ensure it always exists for downstream code.
if (!"judge_skipped" %in% names(raw)) raw$judge_skipped <- NA_character_
if (!"skip_reason"  %in% names(raw)) raw$skip_reason  <- NA_character_

df <- raw |>
  mutate(across(all_of(L2_FEATURES), to_int_binary),
         parse_error   = if_else(parse_error   %in% c("True","TRUE","1"), TRUE, FALSE, missing = FALSE),
         judge_skipped = if_else(judge_skipped %in% c("True","TRUE","1"), TRUE, FALSE, missing = FALSE)) |>
  mutate(model = factor(model, levels = MODEL_ORDER),
         family = factor(MODEL_FAMILY[as.character(model)],
                         levels = c("OpenAI","Anthropic","Google","xAI","Meta","Qwen")))

# Defensive: drop rows that don't match the current prompt SHA256, EXCEPT skips
# (which never reach the judge so don't carry the hash).
df <- df |>
  filter(is.na(judge_prompt_sha256) |
           judge_prompt_sha256 == "" |
           judge_prompt_sha256 == PROMPT_SHA256)

# Add response length features (word, char) on successful judge rows.
df <- df |>
  mutate(
    response_chars = nchar(replace_na(response, "")),
    response_words = stringr::str_count(replace_na(response, ""), "\\S+")
  )

# ---- Service-level refusal handling ----------------------------------------
# A judge_skipped=TRUE row is a target-model API refusal: the user got *no*
# content back at all (or content was blocked upstream). The team's working
# decision: treat these as an "ultra-bare" refusal in L0/L1 views, but filter
# out of any view comparing the *form* of refusals across models (L2 onward)
# so model comparisons stay apples-to-apples.
#
# Two augmented frames:
#   df_full  — all 3200 rows, with judge_skipped given synthetic L0/L1 labels
#              ("Non-compliance" / "Service-Level"). Use for L0/L1 *including*
#              service-level refusals.
#   df_clean — successfully judged rows only (drops 78 skipped + 26 parse
#              errors). Use for L2 work and for "in-voice" L0/L1 views.
df_full <- df |>
  filter(!is.na(model_key)) |>
  mutate(
    layer0 = if_else(judge_skipped & is.na(layer0), "Non-compliance", layer0),
    layer1 = if_else(judge_skipped & is.na(layer1), "Service-Level", layer1),
    refusal_voice = case_when(
      judge_skipped                  ~ "service",   # provider/API blocked output
      layer0 == "Non-compliance"     ~ "model",     # model refused in its own voice
      TRUE                            ~ "compliant"
    )
  )

# Convenience splits.
df_clean <- df |> filter(!parse_error, !is.na(layer0))   # successful judge rows
df_nc    <- df_clean |> filter(layer0 == "Non-compliance")

# ---- Balanced cohort for cross-model comparisons ---------------------------
# A prompt is "clean" iff *every* model has a successfully judged row for it
# (no judge_skipped, no parse_error, non-null layer0). This positive
# formulation is robust to partially-collected models: a prompt absent from
# some model's data is correctly excluded.
N_MODELS <- length(unique(df$model))
BALANCED_IDS <- df |>
  filter(!judge_skipped, !parse_error, !is.na(layer0)) |>
  count(final_sample_id) |>
  filter(n == N_MODELS) |>
  pull(final_sample_id)

ALL_NC_IDS <- df |>
  filter(layer0 == "Non-compliance") |>
  count(final_sample_id) |>
  filter(n == N_MODELS) |>
  pull(final_sample_id)

df_balanced     <- df_clean |> filter(final_sample_id %in% BALANCED_IDS)
df_balanced_nc  <- df_balanced |> filter(layer0 == "Non-compliance")

df_all_nc <- df_clean |> filter(final_sample_id %in% ALL_NC_IDS)

# ---- SorryBench fine-grained sub-categories --------------------------------
# 02_collect_responses didn't propagate sorry_bench_category through to the
# response files, so we join it back from the source prompt file.
sb_lookup <- jsonlite::fromJSON(file.path("..", "data", "prompts", "sampled_200_final_queries.json"),
                                simplifyVector = TRUE) |>
  as_tibble() |>
  filter(source == "sorrybench") |>
  select(final_sample_id, sb_category = sorry_bench_category_name) |>
  mutate(final_sample_id = as.character(final_sample_id))

attach_sb <- function(d) {
  d |>
    mutate(final_sample_id = as.character(final_sample_id)) |>
    left_join(sb_lookup, by = "final_sample_id")
}

df          <- attach_sb(df)
df_full     <- attach_sb(df_full)
df_clean    <- attach_sb(df_clean)
df_nc       <- attach_sb(df_nc)
df_balanced <- attach_sb(df_balanced)
df_balanced_nc <- attach_sb(df_balanced_nc)
```

# Overview

This document characterises the refusal behaviour of `r length(unique(df$model))`
LLMs across `r nrow(df)` prompt-response pairs sampled from LMSYS/WildChat and
SorryBench, judged by GPT-5.5 (non-reasoning) using the codebook-aligned prompt
(SHA256 prefix `r substr(PROMPT_SHA256, 1, 12)`...).

It is descriptive scaffolding for the paper — not the final analysis. Outputs
are organised as:

1. **Coverage** — how many records per model reached the judge, plus the split
   between in-voice refusals and **service-level (API)** refusals.
2. **Service-level refusals** — characterising the upstream-blocked rows as an
   "ultra-bare" refusal class: the user gets no model voice at all.
3. **Layer 0 (Action)** — Full / Partial / Non-compliance distribution.
4. **Layer 1 (Rationale)** — Bare / Capacity / Policy / Ethics (+ Service-level), within NC.
5. **Layer 2 (Form features)** — prevalence by feature, within NC. Restricted
   to a **balanced cohort** of prompts where every model produced a judgeable
   response, so cross-model comparisons are apples-to-apples.
6. **Harm category breakdowns** — Layer 0/1/2 by Llama Guard category.
7. **Cross-cutting analyses** — reasoning vs non-reasoning pairs, model families.
8. **Flavour** — sampled responses per model and category, for qualitative read.

**Cohorts at a glance**:
`df` = all `r nrow(df)` rows · `df_full` = same, with service-level rows
labelled L0=NC / L1=Service-Level · `df_clean` = `r nrow(df_clean)`
successfully judged rows · `df_balanced` = `r nrow(df_balanced)` rows from
`r length(BALANCED_IDS)` prompts where **every** model produced a judgeable
response (apples-to-apples).

---

# 1. Coverage and service-level refusals

## 1.0 Headline numbers for the results-section prose

Three small tables to lift into the opening paragraph and surrounding text.

### Top-line action distribution

Denominator = all `r nrow(df)` in-scope responses. Categories sum to 100%.
"Uncoded" = the judge itself failed to classify (meta-refusal or upstream
classifier block on the judge); disclose in a footnote.

```{r headline-overall}
headline_overall <- df |>
  mutate(category = case_when(
    parse_error                    ~ "Uncoded (judge failure)",
    judge_skipped                  ~ "Service-level (API filter)",
    layer0 == "Full compliance"    ~ "Full compliance",
    layer0 == "Partial compliance" ~ "Partial compliance",
    layer0 == "Non-compliance"     ~ "In-voice non-compliance",
    TRUE                           ~ NA_character_
  )) |>
  count(category) |>
  mutate(
    pct = n / sum(n),
    pct_str = percent(pct, accuracy = 0.1)
  ) |>
  arrange(factor(category, levels = c(
    "Full compliance", "Partial compliance",
    "Service-level (API filter)", "In-voice non-compliance",
    "Uncoded (judge failure)"
  )))

headline_overall |>
  select(category, n, pct_str) |>
  kable(col.names = c("Action category", "n", "% of corpus"),
        caption = "Top-line action distribution across all in-scope responses. Use these numbers in the opening paragraph of Results.") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Most- and least-compliant models

```{r headline-model-ranking}
model_ranking <- df_full |>
  filter(!is.na(layer0)) |>
  group_by(model) |>
  summarise(
    n                = n(),
    pct_full         = mean(layer0 == "Full compliance"),
    pct_partial      = mean(layer0 == "Partial compliance"),
    pct_any_comply   = mean(layer0 %in% c("Full compliance", "Partial compliance")),
    pct_nc           = mean(layer0 == "Non-compliance"),
    pct_service      = mean(judge_skipped, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(pct_any_comply)) |>
  mutate(across(starts_with("pct_"), \(x) percent(x, accuracy = 0.1)))

model_ranking |>
  kable(col.names = c("Model", "n", "Full", "Partial", "Full+Partial",
                      "Non-compliance", "Service-level"),
        caption = "Per-model compliance ranking (sorted by Full+Partial). Service-level rows are included in Non-compliance for df_full.") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Variation in refusal rate by harm category

```{r headline-by-category}
category_ranking <- df_full |>
  filter(!is.na(layer0), !is.na(primary_llama_guard_category_name)) |>
  group_by(primary_llama_guard_category_name) |>
  summarise(
    n           = n(),
    pct_full    = mean(layer0 == "Full compliance"),
    pct_partial = mean(layer0 == "Partial compliance"),
    pct_nc      = mean(layer0 == "Non-compliance"),
    pct_service = mean(judge_skipped, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(pct_nc)) |>
  mutate(across(starts_with("pct_"), \(x) percent(x, accuracy = 0.1)))

category_ranking |>
  kable(col.names = c("Harm category", "n", "Full", "Partial",
                      "Non-compliance", "Service-level"),
        caption = "Action distribution by Llama Guard harm category, sorted by NC rate. Use for the by-category variation sentence in Results.") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Inline values for the prose

For copy-paste into the LaTeX:

- Full compliance: **`r percent(mean(df$layer0 == "Full compliance", na.rm = TRUE), accuracy = 0.1)`**
- Partial compliance: **`r percent(mean(df$layer0 == "Partial compliance", na.rm = TRUE), accuracy = 0.1)`**
- Service-level (API filter): **`r percent(mean(df$judge_skipped), accuracy = 0.1)`**
- In-voice non-compliance: **`r percent(mean(df_full$layer0 == "Non-compliance" & !df_full$judge_skipped & !df_full$parse_error, na.rm = TRUE), accuracy = 0.1)`**
- Uncoded (judge failure, footnote): **`r percent(mean(df$parse_error & !df$judge_skipped), accuracy = 0.1)`**

Most compliant model: **`r model_ranking$model[1]`** (`r model_ranking$pct_any_comply[1]` Full+Partial). Least compliant: **`r model_ranking$model[nrow(model_ranking)]`** (`r model_ranking$pct_any_comply[nrow(model_ranking)]`).

NC range across harm categories: **`r category_ranking$pct_nc[nrow(category_ranking)]`** (`r category_ranking$primary_llama_guard_category_name[nrow(category_ranking)]`) to **`r category_ranking$pct_nc[1]`** (`r category_ranking$primary_llama_guard_category_name[1]`).

## 1.1 Per-model coverage

```{r coverage}
coverage_tbl <- df |>
  group_by(model) |>
  summarise(
    n_total     = n(),
    n_judged    = sum(!parse_error & !judge_skipped & !is.na(layer0)),
    n_service   = sum(judge_skipped),
    n_parse_err = sum(parse_error),
    .groups = "drop"
  ) |>
  mutate(pct_judged  = percent(n_judged  / n_total, accuracy = 0.1),
         pct_service = percent(n_service / n_total, accuracy = 0.1))

coverage_tbl |>
  kable(caption = "Per-model coverage. **n_service** = target-model API refusals (provider blocked output / refusal native_finish_reason). **n_parse_err** = judge meta-refusals (judge model wouldn't classify the content). Both are filtered out of the balanced cohort downstream.") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## 1.2 Service-level (API) refusals — the "ultra-bare" class

When a target model's API returns an empty body, the user receives **no
response at all** — not even a refusal in the model's voice. From an
end-user perspective this is the bluntest possible refusal: there is no
acknowledgement, no rationale, no alternative. We can think of this as a **service-level refusal** for closed models that are tied to a specific service provider. Alternatively,if a query triggers this kind of refusal from *any* model/service, we should remove it from analyses for *all* models. 

### Mechanism breakdown

```{r service-mech}
service_mech <- df |>
  filter(judge_skipped) |>
  group_by(family, model) |>
  summarise(
    n = n(),
    providers = paste(unique(na.omit(provider)), collapse = ","),
    native_finish = paste(names(sort(table(na.omit(native_finish_reason)), decreasing = TRUE))[1:2] |> na.omit(), collapse = ","),
    .groups = "drop"
  ) |>
  arrange(family, desc(n))

service_mech |>
  kable(caption = "Per-model service-level refusal mechanism. **native_finish** shows the top-1/2 finish reasons returned by the underlying provider.") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Service-level refusal rate by family

```{r service-by-family}
df |>
  group_by(family) |>
  summarise(
    n_total = n(),
    n_service = sum(judge_skipped),
    rate = n_service / n_total,
    .groups = "drop"
  ) |>
  mutate(rate_str = percent(rate, accuracy = 0.1)) |>
  arrange(desc(rate)) |>
  kable(caption = "Aggregate service-level refusal rate by model family.") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Where service-level refusals concentrate (by harm category)

```{r service-by-cat, fig.height = 5}
service_cat <- df |>
  filter(judge_skipped) |>
  count(primary_llama_guard_category_name, family) |>
  arrange(desc(n))

ggplot(service_cat,
       aes(reorder(primary_llama_guard_category_name, n), n, fill = family)) +
  geom_col() +
  coord_flip() +
  labs(x = NULL, y = "Service-level refusals",
       title = "Service-level refusals concentrate in a few harm categories",
       subtitle = "Stacked by model family") +
  theme_minimal(base_size = 11)
```

### Examples

```{r}

df |>
  filter(judge_skipped == T) |>
  group_by(query, llama_guard_category_names) |>
  summarize(n_models = n_distinct(model_key)) |>
  arrange(-n_models)

```

### Sonnet 3.7 vs Sonnet 4.6: the Constitutional-Classifier signal

Sonnet 3.7 returned **zero** empty responses; Sonnet 4.6 returned `r sum(df$judge_skipped & df$model == "claude-sonnet-4.6")`.
All Sonnet 4.6 empties carry `native_finish_reason='refusal'`, and a direct
single-row probe against the Anthropic provider on OpenRouter (bypassing
Amazon Bedrock) showed the same outcome — i.e. the blocking is happening at
the Anthropic API itself, consistent with the Constitutional-Classifier-style
input filter Anthropic added with the Claude 4.x line.

```{r service-claude-detail}
df |>
  filter(family == "Anthropic") |>
  group_by(model) |>
  summarise(
    n_service = sum(judge_skipped),
    n_total = n(),
    rate = percent(n_service / n_total, accuracy = 0.1),
    top_category = {
      tab <- table(primary_llama_guard_category_name[judge_skipped])
      if (length(tab) == 0) NA_character_ else names(sort(tab, decreasing = TRUE))[1]
    },
    .groups = "drop"
  ) |>
  kable(caption = "Anthropic line — pre-generation classifier appears to scale with the 4.x rollout.") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# 2. Layer 0: Action distribution

## 2.1 Per-model distribution

The first panel includes **service-level refusals** as a separate category to
show the user-facing reality (no model voice at all). The second panel hides
them and shows the same distribution among in-voice responses only.


```{r l0-by-model-incl, fig.height = 7}
l0_levels_incl <- c("Full compliance", "Partial compliance",
                    "Non-compliance", "API Error")
l0_pal_incl    <- c("Full compliance"    = "#1b9e77",
                    "Partial compliance" = "#ffcc44",
                    "Non-compliance"     = "#cc0022",
                    "API Error"      = "#252525")

l0_tbl_incl <- df_full |>
  filter(!is.na(layer0)) |>
  mutate(layer0_view = case_when(
    refusal_voice == "service" ~ "API Error",
    TRUE                        ~ layer0
  )) |>
  count(model, layer0_view) |>
  group_by(model) |>
  mutate(prop = n / sum(n),
         layer0_view = factor(layer0_view, levels = l0_levels_incl)) |>
  ungroup()

ggplot(l0_tbl_incl, aes(model, prop, fill = layer0_view)) +
  geom_col(width = 0.8) +
  scale_y_continuous(labels = percent_format(1)) +
  scale_fill_manual(values = l0_pal_incl) +
  coord_flip() +
  labs(x = NULL, y = NULL, fill = "Layer 0",
       title = "Action distribution INCLUDING service-level refusals",
       subtitle = "All 200 prompts per model; black = upstream API refusal") +
  theme_minimal(base_size = 12)


 
```

```{r}


ggplot(l0_tbl_incl, aes(model, prop, fill = layer0_view)) +
  geom_col(width = 0.8) +
  scale_y_continuous(labels = percent_format(1)) +
  scale_fill_manual(values = l0_pal_incl) +
  coord_flip() +
  labs(x = NULL, y = NULL, fill = "Layer 0") +
  theme_minimal(base_size = 18) + 
  guides(
    fill = guide_legend(
      nrow = 2,
      byrow = TRUE
    )
  ) +
  theme_minimal(base_size = 18) + 
  theme(
    legend.position = "bottom",
    legend.justification = "left",
    legend.box.just = "left",
    legend.box.margin = margin(t = 8, r = 0, b = 0, l = -95)
  )

ggsave("figs/layer0.pdf")


```

```{r l0-by-model, fig.height = 7}
l0_levels <- c("Full compliance", "Partial compliance", "Non-compliance")
l0_pal    <- c("Full compliance" = "#1b9e77",
               "Partial compliance" = "#d95f02",
               "Non-compliance" = "#7570b3")

l0_tbl <- df_balanced |>
  count(model, layer0) |>
  group_by(model) |>
  mutate(prop = n / sum(n),
         layer0 = factor(layer0, levels = l0_levels)) |>
  ungroup()

ggplot(l0_tbl, aes(model, prop, fill = layer0)) +
  geom_col(width = 0.8) +
  scale_y_continuous(labels = percent_format(1)) +
  scale_fill_manual(values = l0_pal) +
  coord_flip() +
  labs(x = NULL, y = NULL, fill = "Layer 0",
       title = "Action distribution by model",
       subtitle = "Balanced cohort, in-voice only",
       caption = paste0("n = ", length(BALANCED_IDS), " queries")) +
  theme_minimal(base_size = 12)
```

## 2.2 Refusal rate (NC) ranking

Two refusal-rate columns: one over the user-facing total (service-level
refusals count as refusals), one over in-voice judged responses only.

```{r refusal-rate}
df_full |>
  group_by(model) |>
  summarise(
    n_total          = n(),
    n_judged         = sum(!judge_skipped & !parse_error),
    nc_inclusive     = mean(layer0 == "Non-compliance"),
    nc_inv_only      = sum(layer0 == "Non-compliance" & !judge_skipped) /
                       pmax(sum(!judge_skipped & !parse_error), 1),
    service_rate     = mean(judge_skipped),
    .groups = "drop"
  ) |>
  arrange(desc(nc_inclusive)) |>
  mutate(across(c(nc_inclusive, nc_inv_only, service_rate),
                ~percent(.x, accuracy = 0.1))) |>
  kable(caption = "Refusal-rate ranking. nc_inclusive = NC+service over all 200; nc_inv_only = NC over judged-only.") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# 3. Layer 1: Refusal basis (among Non-compliance)

This section shows two parallel views: an **inclusive** L1 distribution where
service-level refusals are added as a fifth class (`Service-Level`), and an
**in-voice** view restricted to refusals where the user got a response.

## 3.0 Headline numbers for the L1 prose

Three tables for the rationale paragraph. Denominator is **in-voice non-compliance only** (service-level rows excluded — those have no model-voice rationale to assess).

### Overall L1 distribution

```{r l1-headline-overall}
l1_headline <- df_nc |>
  count(layer1) |>
  mutate(pct     = n / sum(n),
         pct_str = percent(pct, accuracy = 0.1)) |>
  arrange(desc(n))

l1_headline |>
  select(layer1, n, pct_str) |>
  kable(col.names = c("Refusal basis", "n", "% of in-voice NC"),
        caption = "Layer 1 distribution across in-voice non-compliances. Use these numbers in the L1 paragraph.") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Per-model L1 distribution

```{r l1-headline-per-model}
l1_per_model <- df_nc |>
  group_by(model, layer1) |>
  summarise(n = n(), .groups = "drop_last") |>
  mutate(pct = n / sum(n)) |>
  ungroup()

l1_per_model_wide <- l1_per_model |>
  select(model, layer1, pct) |>
  pivot_wider(names_from = layer1, values_from = pct, values_fill = 0) |>
  mutate(dominant_basis = pmap_chr(
    list(`Bare`, `Capacity-based`, `Ethics-based`, `Policy-based`),
    \(b, c, e, p) {
      vals <- c(Bare = b, `Capacity-based` = c, `Ethics-based` = e, `Policy-based` = p)
      names(vals)[which.max(vals)]
    }
  )) |>
  arrange(desc(`Ethics-based`)) |>
  mutate(across(c(`Bare`, `Capacity-based`, `Ethics-based`, `Policy-based`),
                \(x) percent(x, accuracy = 0.1)))

l1_per_model_wide |>
  select(model, `Ethics-based`, `Bare`, `Policy-based`, `Capacity-based`, dominant_basis) |>
  kable(caption = "Per-model L1 distribution within in-voice NC, sorted by Ethics-based share. `dominant_basis` is the largest category for each model.") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Bare-dominant models

```{r l1-bare-dominant}
bare_ranking <- l1_per_model |>
  filter(layer1 == "Bare") |>
  arrange(desc(pct)) |>
  mutate(pct_str = percent(pct, accuracy = 0.1))

bare_ranking |>
  select(model, n, pct_str) |>
  kable(col.names = c("Model", "n Bare", "% Bare (of NC)"),
        caption = "Models ranked by share of Bare refusals within in-voice NC. Three models cross 50%.") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Inline values for the prose

```{r l1-inline-vals, include = FALSE}
pct_ethics    <- l1_headline$pct_str[l1_headline$layer1 == "Ethics-based"]
pct_bare      <- l1_headline$pct_str[l1_headline$layer1 == "Bare"]
pct_policy    <- l1_headline$pct_str[l1_headline$layer1 == "Policy-based"]
pct_capacity  <- l1_headline$pct_str[l1_headline$layer1 == "Capacity-based"]

top_ethics_row <- l1_per_model_wide |> slice(1)
bare_dom       <- bare_ranking |> filter(pct > 0.5)
```

- Overall **Ethics-based: `r pct_ethics`**
- Overall **Bare: `r pct_bare`**
- Overall **Policy-based: `r pct_policy`**
- Overall **Capacity-based: `r pct_capacity`**
- Top ethics-based model: **`r top_ethics_row$model`** at **`r top_ethics_row[["Ethics-based"]]`**
- Bare-dominant models (>50% Bare): **`r paste(sprintf("%s (%s)", bare_dom$model, bare_dom$pct_str), collapse = "; ")`**

## 3.1 Inclusive view (model voice + service-level)

```{r l1-inclusive, fig.height = 7}
l1_levels_incl <- c("Service-Level", "Bare", "Capacity-based",  "Policy-based","Ethics-based")
l1_pal_incl    <- c("Bare" = "#bdbdbd",
                    "Capacity-based" = "#fdae61",
                    "Policy-based" = "#abd9e9",
                    "Ethics-based" = "#2c7bb6",
                    "Service-Level" = "#252525")

l1_incl_tbl <- df_full |>
  filter(layer0 == "Non-compliance", layer1 %in% l1_levels_incl) |>
  count(model, family, layer1) |>
  group_by(model, family) |>
  mutate(prop = n / sum(n),
         layer1 = factor(layer1, levels = l1_levels_incl)) |>
  ungroup()

ggplot(l1_incl_tbl, aes(model, prop, fill = layer1)) +
  geom_col(width = 0.8) +
  scale_y_continuous(labels = percent_format(1)) +
  scale_fill_manual(values = l1_pal_incl) +
  coord_flip() +
  labs(x = NULL, y = NULL, fill = "Layer 1",
       title = "L1 basis including service-level refusals",
       subtitle = "All Non-compliance — model voice + upstream-blocked",
       caption = "n = 200 queries") +
  theme_minimal(base_size = 12)
```

## 3.2 In-voice view (judged refusals only)

```{r l1-by-model, fig.height = 7}
l1_levels <- c("Bare", "Capacity-based", "Policy-based", "Ethics-based")
l1_pal    <- c("Bare" = "#bdbdbd",
               "Capacity-based" = "#fdae61",
               "Policy-based" = "#abd9e9",
               "Ethics-based" = "#2c7bb6")

l1_tbl <- df_nc |>
  filter(layer1 %in% l1_levels) |>
  count(model, layer1, family) |>
  group_by(model, family) |>
  mutate(prop = n / sum(n),
         layer1 = factor(layer1, levels = l1_levels)) |>
  ungroup()

ggplot(l1_tbl, aes(model, prop, fill = layer1)) +
  geom_col(width = 0.8) +
  scale_y_continuous(labels = percent_format(1)) +
  scale_fill_manual(values = l1_pal) +
  coord_flip() +
  labs(x = NULL, y = NULL, fill = "Layer 1",
       title = "Refusal basis among Non-compliance",
       subtitle = "in-voice refusals",
       caption = paste0("n = ", nrow(df_nc))) +
  theme_minimal(base_size = 12)
```

```{r}

## Paper: Figure 2?

ggplot(l1_tbl, aes(model, prop, fill = layer1)) +
  geom_col(width = 0.8) +
  scale_y_continuous(labels = percent_format(1)) +
  scale_fill_manual(values = l1_pal) +
  coord_flip() +
  labs(x = "Model", y = "Proportion", fill = "Layer 1") +
  theme_minimal(base_size = 18) + 
  theme(
    legend.position = "bottom",
    ) + 
  facet_grid(family ~., scales="free", space="free")

# TODO: Fix facet labeller and move axis left

ggsave("figs/layer1_balanced_cohort.pdf", width=8, height=7)

```

```{r l1-table}
df_nc |>
  filter(layer1 %in% l1_levels) |>
  count(model, layer1) |>
  group_by(model) |>
  mutate(pct = percent(n / sum(n), accuracy = 0.1)) |>
  select(model, layer1, n, pct) |>
  pivot_wider(names_from = layer1, values_from = c(n, pct), values_fill = list(n = 0, pct = "0%")) |>
  kable(caption = paste0("Layer 1 distribution by model")) |>
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 11)
```

## 3.3 Layer 0 and 1 joint figure

I thought a joint view might be more space efficient but there's too much going on.

It also does make me wonder whether we need to do follow-up analyses accounting for the fact that models are refusing different requests when we look at their L0 and L1...

```{r l0-1-joint, fig.height = 7}

l0_1_levels_incl <- c("Full compliance", "Partial compliance",
                    "Service-Level", "Bare", "Capacity-based", "Policy-based", "Ethics-based")

l0_1_pal_incl    <- c("Full compliance"    = "#1b9e77",
                    "Partial compliance" = "#d95f02",
                    "Bare" = "#bdbdbd",
                     "Capacity-based" = "#fdae61",
                     "Policy-based" = "#abd9e9",
                     "Ethics-based" = "#2c7bb6",
                    "Service-Level"      = "#252525")


l0_1_tbl_incl <- df_full |>
  mutate(layer0_1_view = case_when(
    refusal_voice == "service" ~ "Service-Level",
    layer0 == "Non-compliance" ~ layer1,
    TRUE                        ~ layer0
  )) |>
  count(model, layer0_1_view) |>
  group_by(model) |>
  mutate(prop = n / sum(n),
         layer0_1_view = factor(layer0_1_view, levels = l0_1_levels_incl)) |>
  ungroup()

ggplot(l0_1_tbl_incl, aes(model, prop, fill = layer0_1_view)) +
  geom_col(width = 0.8) +
  scale_y_continuous(labels = percent_format(1)) +
  scale_fill_manual(values = l0_1_pal_incl) +
  coord_flip() +
  labs(y = NULL, x = "Model", fill = "Layer 0 & 1") +
  theme_minimal(base_size = 16) + 
    theme(legend.position = "bottom")

```

---

# 4. Layer 2: Form features (among Non-compliance)

## 4.1 Feature prevalence heatmap

```{r l2-heatmap, fig.height = 7}
l2_tbl <- df_nc |>
  select(model, all_of(L2_FEATURES)) |>
  pivot_longer(-model, names_to = "feature", values_to = "value") |>
  filter(!is.na(value)) |>
  group_by(model, feature) |>
  summarise(rate = mean(value), n = n(), .groups = "drop") |>
  mutate(feature = factor(feature, levels = L2_FEATURES))

ggplot(l2_tbl, aes(feature, model, fill = rate)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  geom_text(aes(label = percent(rate, accuracy = 1)), size = 3, vjust = 1) +
  scale_fill_gradient(low = "#f7f7f7", high = "#08519c",
                      labels = percent_format(1), name = "Rate") +
  scale_x_discrete(position = "top") +
  labs(x = NULL, y = NULL,
       title = "L2 feature rates by model (within Non-compliance)") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 0, vjust = 0))
```

```{r}


ggplot(l2_tbl, aes(feature, model, fill = rate)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  geom_text(aes(label = percent(rate, accuracy = 1)), size = 3, vjust = 1) +
  scale_fill_gradient(low = "#f7f7f7", high = "#08519c",
                      labels = percent_format(1), name = "Rate") +
  scale_x_discrete(position = "top") +
  labs(x = NULL, y = NULL,
       title = "L2 feature rates by model (within Non-compliance)") +
  theme_minimal(base_size = 18) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 0, vjust = 0),
    legend.position = "bottom"
    )

library(scales)
library(stringr)
library(grid)

ggplot(l2_tbl, aes(feature, model, fill = rate)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  geom_text(aes(label = percent(rate, accuracy = 1)), size = 3, vjust = 1) +
  scale_fill_gradient(
    low = "#f7f7f7",
    high = "#08519c",
    labels = percent_format(1),
    breaks = seq(0, 1, .25),
    name = "Rate",
    guide = guide_colourbar(
      title.position = "left",
      title.vjust = 0.5,
      barwidth = unit(4.5, "cm"),
      barheight = unit(0.45, "cm")
    )
  ) +
  scale_x_discrete(
    position = "top",
    labels = function(x) str_wrap(x, width = 14)
  ) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 18) +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 0,
      vjust = 0,
      lineheight = 0.85
    ),
    legend.position = "bottom",
    legend.justification = "center",
    legend.box = "horizontal",
    legend.margin = margin(t = 8),
    plot.margin = margin(t = 35, r = 35, b = 10, l = 10)
  )

# ggsave("figs/layer2_model_heatmap.pdf", width = 12, height = 9, dpi = 300)
```

Takehomes:
  
  - Most models are very explicit, but the Qwens tend to be more implicit
  - Some models never apologise (llama, later Claude & GPT), some do it occasionally (Qwen, Grok, Sonnet 3.7, GPT-5) others do it all the time (early GPT)
  - Models never hedge
  - Most models don't preface or positively align. Some do it occasionally (Gemini, Claude Sonnet 3.7)
  - Solidarity is rare for most models, but moderately common for some (Gemini > Opus 4.7 > Qwen ~ Sonnets ~ GPT-5)
  - Models vary a lot wrt negative stance. Some do it a lot (Qwens, Grok, Gemini, Claude), some fairly often (GPT5+, LLama 3)
  
  Overall: Gemini is v interesting. Lots of aligning/affiliative features but also lots of negative stance, Should find examples
  
Follow up q's:
  - Why are models (e.g. qwen) implicit vs explicit? What predicts it? Harm category?
  - What correlates with apologies for "sometimes apologisers"?
  

### Implicit vs Explicit (Claude vs Qwen)
  

```{r}

qwen_implicit_sample <- 
  df_nc |>
    filter(model_key == "qwen3-32b-reasoning", implicit_nc == 1) |>
    select(llama_guard_category_names, query, response)

opus_explicit_sample <- 
  df_nc |>
    filter(model_key == "claude-opus-4.6-reasoning", explicit_nc == 1) |>
    select(llama_guard_category_names, query, response)

merge(qwen_implicit_sample, opus_explicit_sample, by=c("query", "llama_guard_category_names")) |>
  rename("Qwen Response" = "response.x", "Claude Response" = "response.y") 

```

## 4.2 Feature-by-feature ranking

```{r l2-ranking}
l2_tbl |>
  group_by(feature) |>
  summarise(
    overall = percent(weighted.mean(rate, n), accuracy = 0.1),
    top_model = model[which.max(rate)],
    top_rate  = percent(max(rate), accuracy = 0.1),
    bot_model = model[which.min(rate)],
    bot_rate  = percent(min(rate), accuracy = 0.1),
    .groups = "drop"
  ) |>
  kable(caption = "Where each L2 feature peaks and bottoms out.") |>
  kable_styling(bootstrap_options = c("striped","hover"), font_size = 11)
```

---

# 5. By harm category

## 5.1 L0 by category × model (inclusive of service-level)

This view treats service-level refusals as their own category so we can see
*where* upstream blocking concentrates per family.

```{r l0-by-cat, fig.height = 9}
df_full |>
  mutate(layer0_view = if_else(refusal_voice == "service", "Service-Level", layer0)) |>
  filter(!is.na(layer0_view) & layer0_view != "") |>
  count(primary_llama_guard_category_name, model, layer0_view) |>
  group_by(primary_llama_guard_category_name, model) |>
  mutate(prop = n / sum(n),
         layer0_view = factor(layer0_view, levels = l0_levels_incl)) |>
  ungroup() |>
  ggplot(aes(model, prop, fill = layer0_view)) +
    geom_col(width = 0.8) +
    coord_flip() +
    facet_wrap(~ primary_llama_guard_category_name, ncol = 3) +
    scale_y_continuous(labels = percent_format(1), breaks = c(0, 0.5, 1)) +
    scale_fill_manual(values = l0_pal_incl) +
    labs(x = NULL, y = NULL, fill = "Layer 0",
         title = "Action distribution: model × harm category (incl. service-level)") +
    theme_minimal(base_size = 10) +
    theme(legend.position = "bottom",
          strip.text = element_text(face = "bold", size = 9))
```

## 5.2 Refusal rate by category

Two rates: inclusive (model-voice refusal OR upstream block) and in-voice only.

```{r refusal-rate-by-cat}
df_full |>
  group_by(primary_llama_guard_category_name) |>
  summarise(
    n_total           = n(),
    refusal_inclusive = mean(layer0 == "Non-compliance"),       # model-voice + service-level
    refusal_in_voice  = sum(layer0 == "Non-compliance" & !judge_skipped) /
                        pmax(sum(!judge_skipped & !parse_error), 1),
    service_rate      = mean(judge_skipped),
    sd_across_models  = sd(tapply(layer0 == "Non-compliance", model, mean), na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(refusal_inclusive)) |>
  mutate(across(c(refusal_inclusive, refusal_in_voice, service_rate, sd_across_models),
                ~percent(.x, accuracy = 0.1))) |>
  kable(caption = "Harm categories ranked by inclusive refusal rate. sd_across_models shows between-model variability in the *inclusive* rate.") |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

## 5.3 L1 rationale shifts across categories

Balanced cohort, in-voice only — service-level cases are visualised in §1.2
above.

```{r l1-by-cat}
df_balanced_nc |>
  filter(layer1 %in% l1_levels) |>
  count(primary_llama_guard_category_name, layer1) |>
  group_by(primary_llama_guard_category_name) |>
  mutate(prop = n / sum(n),
         layer1 = factor(layer1, levels = l1_levels)) |>
  ungroup() |>
  ggplot(aes(primary_llama_guard_category_name, prop, fill = layer1)) +
    geom_col(width = 0.8) +
    scale_y_continuous(labels = percent_format(1)) +
    scale_fill_manual(values = l1_pal) +
    coord_flip() +
    labs(x = NULL, y = NULL, fill = "Layer 1",
         title = "L1 rationale by harm category (balanced cohort, in-voice)") +
    theme_minimal(base_size = 11)
```

## 5.4 L2 features by category (selected)

Balanced cohort. Which categories elicit hedging, solidarity, and normative
suggestion?

```{r l2-by-cat-selected, fig.height = 7}
selected_features <- c("apology", "solidarity", "negative_stance",
                       "normative_suggestion", "alternative_offer", "executed_alternative")

df_balanced_nc |>
  select(primary_llama_guard_category_name, all_of(selected_features)) |>
  pivot_longer(-primary_llama_guard_category_name, names_to = "feature", values_to = "value") |>
  filter(!is.na(value)) |>
  group_by(primary_llama_guard_category_name, feature) |>
  summarise(rate = mean(value), .groups = "drop") |>
  mutate(feature = factor(feature, levels = selected_features)) |>
  ggplot(aes(feature, primary_llama_guard_category_name, fill = rate)) +
    geom_tile(colour = "white", linewidth = 0.3) +
    geom_text(aes(label = percent(rate, accuracy = 1)), size = 2.8, vjust = 1) +
    scale_fill_gradient(low = "#f7f7f7", high = "#08519c",
                        labels = percent_format(1)) +
    labs(x = NULL, y = NULL, fill = "Rate",
         title = "Selected L2 features × harm category (balanced)") +
    theme_minimal(base_size = 10) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## 5.5 Full model × L2 feature × harm category

Every cell is the rate of a given L2 feature among that model's NC responses
within that harm category. Cells with fewer than `MIN_CELL_N` NC observations
are masked to grey, because rates over very small denominators are not
meaningful. Split across two figures to stay legible.

```{r l2-mod-cat-prep}
MIN_CELL_N <- 3

l2_mod_cat <- df_balanced_nc |>
  select(model,
         category = primary_llama_guard_category_name,
         all_of(L2_FEATURES)) |>
  pivot_longer(-c(model, category), names_to = "feature", values_to = "value") |>
  filter(!is.na(value)) |>
  group_by(model, category, feature) |>
  summarise(rate = mean(value), n = n(), .groups = "drop") |>
  mutate(rate = if_else(n < MIN_CELL_N, NA_real_, rate),
         feature = factor(feature, levels = L2_FEATURES))

# Order categories by their pooled refusal rate (high → low) so the most
# refusal-heavy categories appear first in the split.
cat_order <- df_balanced |>
  group_by(primary_llama_guard_category_name) |>
  summarise(rr = mean(layer0 == "Non-compliance"), .groups = "drop") |>
  arrange(desc(rr)) |>
  pull(primary_llama_guard_category_name)

l2_mod_cat <- l2_mod_cat |>
  mutate(category = factor(category, levels = cat_order))

cat_part1 <- cat_order[1:7]
cat_part2 <- cat_order[8:length(cat_order)]
```

```{r l2-mod-cat-plot-1, fig.height = 18, fig.width = 11}
ggplot(filter(l2_mod_cat, category %in% cat_part1),
       aes(feature, model, fill = rate)) +
  geom_tile(colour = "white", linewidth = 0.2) +
  geom_text(aes(label = if_else(is.na(rate), "", percent(rate, accuracy = 1))),
            size = 2.4, vjust = 1) +
  facet_wrap(~ category, ncol = 1) +
  scale_fill_gradient(low = "#f7f7f7", high = "#08519c",
                      na.value = "grey92",
                      labels = percent_format(1), name = "Rate") +
  scale_x_discrete(position = "top") +
  labs(x = NULL, y = NULL,
       title = "L2 feature rates: model × harm category (part 1 of 2)",
       subtitle = paste0("Cells with n < ", MIN_CELL_N,
                         " NC responses are masked. Ordered by category refusal rate.")) +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, hjust = 0, vjust = -0.2),
        strip.text = element_text(face = "bold"),
        panel.spacing.y = unit(1.2, "lines"))
```

```{r l2-mod-cat-plot-2, fig.height = 18, fig.width = 11}
ggplot(filter(l2_mod_cat, category %in% cat_part2),
       aes(feature, model, fill = rate)) +
  geom_tile(colour = "white", linewidth = 0.2) +
  geom_text(aes(label = if_else(is.na(rate), "", percent(rate, accuracy = 1))),
            size = 2.4, vjust = 1) +
  facet_wrap(~ category, ncol = 1) +
  scale_fill_gradient(low = "#f7f7f7", high = "#08519c",
                      na.value = "grey92",
                      labels = percent_format(1), name = "Rate") +
  scale_x_discrete(position = "top") +
  labs(x = NULL, y = NULL,
       title = "L2 feature rates: model × harm category (part 2 of 2)",
       subtitle = paste0("Cells with n < ", MIN_CELL_N,
                         " NC responses are masked.")) +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, hjust = 0, vjust = -0.2),
        strip.text = element_text(face = "bold"),
        panel.spacing.y = unit(1.2, "lines"))
```

## 5.6 Richer visualization ideas

The full model × category × feature heatmap is information-dense but visually
noisy. The views below are prototypes for higher-signal alternatives All use the balanced cohort.

### 5.6.1 Aggregated feature families

The 11 substantive L2 features (excluding `explicit_nc`/`implicit_nc` — see
note below) collapse into four interpretable bundles. A response scores 1 on
a bundle if **any** of its constituent features is 1 (OR rule); the bundle
rate is then the proportion of refusals with at least one of the features
present.

```{r feature-families-config}
FEATURE_FAMILY <- list(
  rapport    = c("solidarity", "positive_alignment"),
  concessive = c("apology", "hedge", "explanatory_preface", "role_based"),
  normative  = c("normative_suggestion", "negative_stance", "statement_of_principle"),
  helpful    = c("alternative_offer", "executed_alternative")
)

FEATURE_FAMILY_DESC <- tribble(
  ~bundle,       ~gloss,                                                              ~features,
  "rapport",     "Rapport-building — explicitly aligns with or stands with the user.", paste(FEATURE_FAMILY$rapport,    collapse = ", "),
  "concessive",  "Concessive / softening — yields ground, hedges, or distances the refusal.", paste(FEATURE_FAMILY$concessive, collapse = ", "),
  "normative",   "Normative / position-taking — asserts what the model takes to be right.",   paste(FEATURE_FAMILY$normative,  collapse = ", "),
  "helpful",     "Substitution / redirection — what the model offers instead.",        paste(FEATURE_FAMILY$helpful,    collapse = ", ")
)

FEATURE_FAMILY_DESC |>
  kable(caption = "Four L2-feature bundles used in 5.6.1 and 5.6.3.") |>
  kable_styling(bootstrap_options = c("striped","hover"), font_size = 11, full_width = FALSE)
```

`explicit_nc` and `implicit_nc` are mutually exclusive surface forms of the
refusal act itself (they sum to 1 across NC responses), so they're shown
separately rather than folded into a bundle.

```{r feature-families-plot, fig.height = 7, fig.width = 12}
family_long <- df_balanced_nc |>
  select(model, family, all_of(L2_FEATURES)) |>
  mutate(across(all_of(L2_FEATURES), as.numeric)) |>
  rowwise() |>
  mutate(
    rapport    = as.integer(any(c_across(all_of(FEATURE_FAMILY$rapport))    == 1, na.rm = TRUE)),
    concessive = as.integer(any(c_across(all_of(FEATURE_FAMILY$concessive)) == 1, na.rm = TRUE)),
    normative  = as.integer(any(c_across(all_of(FEATURE_FAMILY$normative))  == 1, na.rm = TRUE)),
    helpful    = as.integer(any(c_across(all_of(FEATURE_FAMILY$helpful))    == 1, na.rm = TRUE))
  ) |>
  ungroup() |>
  select(model, family, rapport, concessive, normative, helpful) |>
  pivot_longer(-c(model, family), names_to = "feature_family", values_to = "value") |>
  group_by(model, family, feature_family) |>
  summarise(rate = mean(value, na.rm = TRUE), .groups = "drop") |>
  mutate(feature_family = factor(feature_family,
                                 levels = c("rapport","concessive","normative","helpful")))

# Within-family release rank from MODEL_ORDER (which is already family→oldest→newest)
release_rank <- tibble(model = MODEL_ORDER) |>
  mutate(family = factor(MODEL_FAMILY[model],
                         levels = c("OpenAI","Anthropic","Google","xAI","Meta","Qwen"))) |>
  group_by(family) |>
  mutate(release_rank = row_number()) |>
  ungroup()

family_long_ranked <- family_long |>
  inner_join(select(release_rank, model, release_rank), by = "model") |>
  mutate(release_rank = factor(release_rank))

# Per-cell model name labels (one per facet cell)
cell_labels <- release_rank |>
  filter(model %in% unique(family_long$model)) |>
  mutate(release_rank = factor(release_rank))

ggplot(family_long_ranked, aes(feature_family, rate, fill = feature_family)) +
  geom_col(width = 0.85) +
  geom_text(
    data = cell_labels,
    aes(x = 3, y = 1.05, label = model),
    inherit.aes = FALSE,
    size = 2.4, fontface = "bold", color = "grey20"
  ) +
  facet_grid(family ~ release_rank,
             scales = "free_x", space = "free_x", switch = "y") +
  scale_y_continuous(labels = percent_format(1),
                     limits = c(0, 1.18),
                     expand = expansion(mult = c(0, 0))) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = NULL, y = NULL, fill = NULL,
       title = "Aggregated L2 feature families by model",
       subtitle = "Rows = company family · columns = within-family release order (1 = oldest)",
       caption = paste0("n = ", length(BALANCED_IDS), " queries (balanced cohort, in-voice refusals)")) +
  theme_minimal(base_size = 9) +
  theme(
    axis.text.x  = element_text(angle = 60, hjust = 1, vjust = 1, size = 7),
    strip.text.y.left = element_text(angle = 0, face = "bold", hjust = 1, size = 9),
    strip.text.x = element_text(face = "bold", size = 8),
    strip.placement = "outside",
    panel.spacing.x = unit(0.2, "lines"),
    panel.spacing.y = unit(0.3, "lines"),
    legend.position = "none"
  )
```

### 5.6.2 Harm severity tiers

The 14 Llama Guard categories collapse into three rough severity tiers. These
are a first-pass assignment — should be aligned across the team on canonical
groupings before any claims rest on them.

```{r severity-tiers-config}
HARM_TIER <- c(
  # high: third-party harm at scale
  "Child Sexual Exploitation" = "high",
  "Indiscriminate Weapons"    = "high",
  "Violent Crimes"            = "high",
  "Hate"                      = "high",
  "Sex-Related Crimes"        = "high",
  # med: personal / contextual harm
  "Defamation"                = "med",
  "Privacy"                   = "med",
  "Sexual Content"            = "med",
  "Non-Violent Crimes"        = "med",
  # low: regulatory / specialised-domain
  "Specialized Advice"        = "low",
  "Intellectual Property"     = "low",
  "Code Interpreter Abuse"    = "low",
  "Elections"                 = "low",
  # self-harm: kept separate — different normative expectations (safeguarding
  # the user, not preventing third-party harm)
  "Suicide & Self-Harm"       = "self-harm"
)

TIER_LEVELS <- c("high","med","low","self-harm")
TIER_PALETTE <- c(high = "#7570b3", med = "#fdae61", low = "#1b9e77", `self-harm` = "#d62728")

HARM_TIER_DESC <- tribble(
  ~tier,        ~gloss,                                                                          ~categories,
  "high",       "Third-party harm at scale; near-universal refusal.",                            paste(names(HARM_TIER)[HARM_TIER == "high"],      collapse = ", "),
  "med",        "Personal / contextual harm; refusal less universal.",                           paste(names(HARM_TIER)[HARM_TIER == "med"],       collapse = ", "),
  "low",        "Regulatory / specialised-domain; refusal often capacity-based.",                paste(names(HARM_TIER)[HARM_TIER == "low"],       collapse = ", "),
  "self-harm",  "Safeguards directed at the user themselves; distinct normative expectations.",  paste(names(HARM_TIER)[HARM_TIER == "self-harm"], collapse = ", ")
)

HARM_TIER_DESC |>
  kable(caption = "Provisional severity tiers used in 5.6.2 and 5.6.3.") |>
  kable_styling(bootstrap_options = c("striped","hover"), font_size = 11, full_width = FALSE)
```

```{r severity-tiers-plot, fig.height = 4.5}
df_balanced_nc |>
  mutate(tier = HARM_TIER[primary_llama_guard_category_name] |> factor(levels = TIER_LEVELS)) |>
  filter(!is.na(tier)) |>
  select(model, family, tier, all_of(L2_FEATURES)) |>
  pivot_longer(all_of(L2_FEATURES), names_to = "feature", values_to = "value") |>
  filter(!is.na(value)) |>
  group_by(tier, feature) |>
  summarise(rate = mean(value), n = n(), .groups = "drop") |>
  mutate(feature = factor(feature, levels = L2_FEATURES)) |>
  ggplot(aes(feature, rate, fill = tier)) +
    geom_col(position = "dodge") +
    scale_y_continuous(labels = percent_format(1)) +
    scale_fill_manual(values = TIER_PALETTE) +
    labs(x = NULL, y = "Rate", fill = "Severity tier",
         title = "L2 feature use shifts with harm severity",
         subtitle = "Pooled across models",
         caption = paste0("n = ", length(BALANCED_IDS), " queries")) +
    theme_minimal(base_size = 10) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

### 5.6.3 Putting it together: model × harm tier × feature bundle

The previous two views each compress one dimension (5.6.1 hides harm; 5.6.2
hides model). Combined: each cell is a model's mean rate on one feature
bundle within one severity tier — the cleanest view of how refusal *form*
varies as both the actor (model) and the target (harm severity) change.

```{r model-tier-bundle, fig.height = 7.5}
mtb <- df_balanced_nc |>
  mutate(tier = HARM_TIER[primary_llama_guard_category_name] |>
                factor(levels = TIER_LEVELS)) |>
  filter(!is.na(tier)) |>
  mutate(across(all_of(L2_FEATURES), as.numeric)) |>
  rowwise() |>
  mutate(
    rapport    = as.integer(any(c_across(all_of(FEATURE_FAMILY$rapport))    == 1, na.rm = TRUE)),
    concessive = as.integer(any(c_across(all_of(FEATURE_FAMILY$concessive)) == 1, na.rm = TRUE)),
    normative  = as.integer(any(c_across(all_of(FEATURE_FAMILY$normative))  == 1, na.rm = TRUE)),
    helpful    = as.integer(any(c_across(all_of(FEATURE_FAMILY$helpful))    == 1, na.rm = TRUE))
  ) |>
  ungroup() |>
  select(model, tier, rapport, concessive, normative, helpful) |>
  pivot_longer(c(rapport, concessive, normative, helpful),
               names_to = "feature_family", values_to = "value") |>
  group_by(model, tier, feature_family) |>
  summarise(rate = mean(value, na.rm = TRUE), n = n(), .groups = "drop") |>
  filter(n >= 3) |>
  mutate(feature_family = factor(feature_family,
                                 levels = c("rapport","concessive","normative","helpful")))

ggplot(mtb, aes(feature_family, model, fill = rate)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  geom_text(aes(label = percent(rate, accuracy = 1)), size = 2.6, vjust = 1) +
  facet_wrap(~ tier, nrow = 1,
             labeller = labeller(tier = function(x) paste0(toupper(substring(x,1,1)),
                                                           substring(x,2)))) +
  scale_fill_gradient(low = "#f7f7f7", high = "#08519c",
                      labels = percent_format(1), name = "Rate") +
  scale_x_discrete(position = "top") +
  labs(x = NULL, y = NULL,
       title = "Model × harm severity × L2 feature bundle",
       subtitle = "Cells: mean bundle rate among NC responses (n ≥ 3 required)",
       caption = paste0("n = ", length(BALANCED_IDS), " queries")) +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 30, hjust = 0, vjust = -0.2),
        strip.text = element_text(face = "bold"))
```

### 5.6.4 Full granularity: model × harm category × feature bundle

Same logic as 5.6.3 but with the 14 individual harm categories instead of
severity tiers. Four panels, one per bundle; rows = model, cols = harm
category (ordered by pooled refusal rate, high → low).

```{r mod-cat-bundle, fig.height = 14, fig.width = 11}
mcb <- df_balanced_nc |>
  mutate(across(all_of(L2_FEATURES), as.numeric)) |>
  rowwise() |>
  mutate(
    rapport    = as.integer(any(c_across(all_of(FEATURE_FAMILY$rapport))    == 1, na.rm = TRUE)),
    concessive = as.integer(any(c_across(all_of(FEATURE_FAMILY$concessive)) == 1, na.rm = TRUE)),
    normative  = as.integer(any(c_across(all_of(FEATURE_FAMILY$normative))  == 1, na.rm = TRUE)),
    helpful    = as.integer(any(c_across(all_of(FEATURE_FAMILY$helpful))    == 1, na.rm = TRUE))
  ) |>
  ungroup() |>
  select(model,
         category = primary_llama_guard_category_name,
         rapport, concessive, normative, helpful) |>
  pivot_longer(c(rapport, concessive, normative, helpful),
               names_to = "feature_family", values_to = "value") |>
  group_by(model, category, feature_family) |>
  summarise(rate = mean(value, na.rm = TRUE), n = n(), .groups = "drop") |>
  filter(n >= MIN_CELL_N) |>
  mutate(feature_family = factor(feature_family,
                                 levels = c("rapport","concessive","normative","helpful")),
         category       = factor(category, levels = cat_order))

ggplot(mcb, aes(category, model, fill = rate)) +
  geom_tile(colour = "white", linewidth = 0.25) +
  geom_text(aes(label = if_else(is.na(rate), "", percent(rate, accuracy = 1))),
            size = 2.4, vjust = 1) +
  facet_wrap(~ feature_family, ncol = 1) +
  scale_fill_gradient(low = "#f7f7f7", high = "#08519c",
                      na.value = "grey92",
                      labels = percent_format(1), name = "Rate") +
  scale_x_discrete(position = "top") +
  labs(x = NULL, y = NULL,
       title = "Model × harm category × feature bundle",
       subtitle = paste0("OR rule per bundle; cells with n < ", MIN_CELL_N, " NC responses masked"),
       caption = paste0("n = ", length(BALANCED_IDS), " queries")) +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, hjust = 0, vjust = -0.2),
        strip.text  = element_text(face = "bold"),
        panel.spacing.y = unit(1.0, "lines"))
```

### 5.6.5 Top / bottom model × harm-category per L2 feature

For each feature, who's at the extremes? Useful for narrative.

```{r extremes-table}
extremes_mod <- df_balanced_nc |>
  select(model, all_of(L2_FEATURES)) |>
  pivot_longer(-model, names_to = "feature", values_to = "value") |>
  filter(!is.na(value)) |>
  group_by(model, feature) |>
  summarise(rate = mean(value), n = n(), .groups = "drop") |>
  filter(n >= 5) |>
  group_by(feature) |>
  summarise(
    top_model = model[which.max(rate)],
    top_rate  = percent(max(rate), accuracy = 0.1),
    bot_model = model[which.min(rate)],
    bot_rate  = percent(min(rate), accuracy = 0.1),
    .groups = "drop"
  )

extremes_cat <- df_balanced_nc |>
  select(category = primary_llama_guard_category_name, all_of(L2_FEATURES)) |>
  pivot_longer(-category, names_to = "feature", values_to = "value") |>
  filter(!is.na(value)) |>
  group_by(category, feature) |>
  summarise(rate = mean(value), n = n(), .groups = "drop") |>
  filter(n >= 5) |>
  group_by(feature) |>
  summarise(
    top_category = category[which.max(rate)],
    top_cat_rate = percent(max(rate), accuracy = 0.1),
    bot_category = category[which.min(rate)],
    bot_cat_rate = percent(min(rate), accuracy = 0.1),
    .groups = "drop"
  )

extremes_mod |>
  left_join(extremes_cat, by = "feature") |>
  mutate(feature = factor(feature, levels = L2_FEATURES)) |>
  arrange(feature) |>
  kable(caption = "Extremes per L2 feature: which model and which harm category push it highest/lowest. (Cells with n<5 excluded.)") |>
  kable_styling(bootstrap_options = c("striped","hover"), font_size = 10)
```

### 5.6.6 Cross-model entropy: which features divide the field?

For each feature, treat model-level rates as a distribution and compute its
**spread** (SD across models). High SD → models disagree on whether to use it.
Low SD → near-universal convention.

```{r feature-entropy, fig.height = 5}
entropy_tbl <- df_balanced_nc |>
  select(model, all_of(L2_FEATURES)) |>
  pivot_longer(-model, names_to = "feature", values_to = "value") |>
  filter(!is.na(value)) |>
  group_by(model, feature) |>
  summarise(rate = mean(value), .groups = "drop") |>
  group_by(feature) |>
  summarise(
    mean_rate = mean(rate),
    sd_rate   = sd(rate),
    range     = diff(range(rate)),
    .groups = "drop"
  ) |>
  arrange(desc(sd_rate)) |>
  mutate(feature = factor(feature, levels = feature))

ggplot(entropy_tbl, aes(feature, sd_rate)) +
  geom_col(fill = "#7570b3") +
  geom_point(aes(y = mean_rate), color = "#d95f02", size = 3) +
  scale_y_continuous(labels = percent_format(1)) +
  coord_flip() +
  labs(x = NULL, y = "SD of model-level rate (bar) / Mean rate (orange dot)",
       title = "Which L2 features divide the field vs. are conventional?",
       subtitle = "Higher bar = more between-model disagreement on whether to use the feature.") +
  theme_minimal(base_size = 11)
```

### 5.6.7 Surprising cells: stance × category and affiliative × category

Two specific "what would be weird?" lookups:

* **Negative stance** is near-universal almost everywhere — where is it *least* deployed?
* **Solidarity** is rare overall — where is it *most* deployed, including in unusual categories like CSE?

```{r surprising-cells-prep}
specific_views <- df_balanced_nc |>
  group_by(model, category = primary_llama_guard_category_name) |>
  summarise(
    n = n(),
    negative_stance_rate    = mean(negative_stance, na.rm = TRUE),
    solidarity_rate         = mean(solidarity, na.rm = TRUE),
    positive_alignment_rate = mean(positive_alignment, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(n >= 3)
```

**Lowest `negative_stance` cells** — where models are oddly *not* condemnatory:

```{r surprising-cells-low-stance}
specific_views |>
  arrange(negative_stance_rate) |>
  slice_head(n = 8) |>
  mutate(across(ends_with("_rate"), ~percent(.x, accuracy = 1))) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped","hover"), font_size = 10)
```

**Highest `solidarity` cells** — where models lean affiliative:

```{r surprising-cells-high-solidarity}
specific_views |>
  arrange(desc(solidarity_rate)) |>
  slice_head(n = 8) |>
  mutate(across(ends_with("_rate"), ~percent(.x, accuracy = 1))) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped","hover"), font_size = 10)
```

### 5.6.8 SorryBench fine-grained categories (SB queries only)

SorryBench provides much finer categorisation (37 sub-categories) than Llama
Guard's 14. Restricting to the 132 SB prompts, here are the categories where
the *inclusive* refusal rate is highest and lowest, and which models disagree
most. Categories with fewer than 3 prompts are kept for completeness but
should be read with caution.

```{r sb-refusal-rate}
df_full |>
  filter(!is.na(sb_category)) |>
  group_by(sb_category) |>
  summarise(
    n_prompts   = n_distinct(final_sample_id),
    n_responses = n(),
    refusal_inclusive = mean(layer0 == "Non-compliance"),
    service_rate      = mean(judge_skipped),
    sd_across_models  = sd(tapply(layer0 == "Non-compliance", model, mean), na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(refusal_inclusive)) |>
  mutate(across(c(refusal_inclusive, service_rate, sd_across_models),
                ~percent(.x, accuracy = 0.1))) |>
  kable(caption = "SorryBench sub-categories ranked by inclusive refusal rate.") |>
  kable_styling(bootstrap_options = c("striped","hover"), font_size = 10)
```

```{r sb-l1, fig.height = 9}
df_balanced_nc |>
  filter(!is.na(sb_category), layer1 %in% l1_levels) |>
  count(sb_category, layer1) |>
  group_by(sb_category) |>
  mutate(prop = n / sum(n),
         layer1 = factor(layer1, levels = l1_levels),
         total_n = sum(n)) |>
  ungroup() |>
  filter(total_n >= 8) |>  # drop sparsely-populated sub-categories from the plot
  ggplot(aes(reorder(sb_category, total_n), prop, fill = layer1)) +
    geom_col(width = 0.85) +
    scale_y_continuous(labels = percent_format(1)) +
    scale_fill_manual(values = l1_pal) +
    coord_flip() +
    labs(x = NULL, y = NULL, fill = "Layer 1",
         title = "L1 rationale by SorryBench sub-category",
         subtitle = "Balanced cohort; sub-categories with <8 in-voice refusals omitted.") +
    theme_minimal(base_size = 10)
```

### 5.6.9 Response length as a basic feature

How long are model responses? Word counts on full-text responses (any L0).

```{r word-counts, fig.height = 5}
df_clean |>
  filter(!is.na(response_words), response_words > 0) |>
  ggplot(aes(model, response_words, fill = family)) +
    geom_boxplot(outlier.size = 0.5, outlier.alpha = 0.4, alpha = 0.7) +
    scale_y_log10() +
    coord_flip() +
    labs(x = NULL, y = "Response length (words, log scale)", fill = "Family",
         title = "Response length by model",
         subtitle = "Across all action types — refusals tend to be shorter than compliances") +
    theme_minimal(base_size = 11)
```

```{r word-counts-by-l0, fig.height = 5}
df_clean |>
  filter(!is.na(response_words), response_words > 0) |>
  group_by(model, layer0) |>
  summarise(median_words = median(response_words),
            q25 = quantile(response_words, 0.25),
            q75 = quantile(response_words, 0.75),
            .groups = "drop") |>
  ggplot(aes(model, median_words, color = layer0)) +
    geom_pointrange(aes(ymin = q25, ymax = q75), position = position_dodge(width = 0.5)) +
    scale_color_manual(values = l0_pal) +
    coord_flip() +
    labs(x = NULL, y = "Median words (IQR)", color = "Layer 0",
         title = "Response length by L0 action and model") +
    theme_minimal(base_size = 11)
```

---

# 6. Cross-cutting analyses

```{r ci-helpers}
# Wilson-score CI for a proportion (more sensible than normal-approx at the
# extremes). Avoids the need for a bootstrap loop for binary L2 rates.
wilson_ci <- function(k, n, z = 1.96) {
  if (n == 0) return(c(estimate = NA_real_, lower = NA_real_, upper = NA_real_))
  p <- k / n
  denom <- 1 + z^2 / n
  centre <- (p + z^2 / (2 * n)) / denom
  half <- z * sqrt(p * (1 - p) / n + z^2 / (4 * n^2)) / denom
  c(estimate = p, lower = max(0, centre - half), upper = min(1, centre + half))
}

# Vectorised wrapper that returns a tibble.
prop_ci <- function(x) {
  x <- x[!is.na(x)]
  ci <- wilson_ci(sum(x == 1), length(x))
  tibble(rate = ci["estimate"], lo = ci["lower"], hi = ci["upper"], n = length(x))
}
```

## 6.1 Reasoning vs non-reasoning

For models with both a base and a reasoning variant: does turning on the
thinking trace change the refusal profile?

```{r reasoning-pairs}
pair_tbl <- df_clean |>
  filter(model %in% c(REASONING_PAIRS$base, REASONING_PAIRS$reasoning)) |>
  mutate(variant = if_else(model %in% REASONING_PAIRS$reasoning, "reasoning", "base"),
         base_model = case_when(
           model %in% REASONING_PAIRS$base ~ as.character(model),
           model %in% REASONING_PAIRS$reasoning ~ str_replace(as.character(model), "-reasoning$", ""),
           TRUE ~ NA_character_
         )) |>
  group_by(base_model, variant) |>
  summarise(
    refusal_rate = mean(layer0 == "Non-compliance"),
    partial_rate = mean(layer0 == "Partial compliance"),
    .groups = "drop"
  ) |>
  pivot_longer(ends_with("_rate"), names_to = "metric", values_to = "rate") |>
  pivot_wider(names_from = variant, values_from = rate) |>
  mutate(delta = reasoning - base)

pair_tbl |>
  mutate(across(c(base, reasoning, delta), ~percent(.x, accuracy = 0.1))) |>
  kable(caption = "Reasoning vs base: Δ in refusal/partial rate. Positive Δ means reasoning variant refuses more.") |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

### L2 feature deltas with CIs (reasoning − base)

Each base/reasoning pair on the balanced cohort, with Wilson 95% CIs on each
side and an overlaid cross-pair mean per feature.

```{r reasoning-l2-deltas, fig.height = 6}
reasoning_long <- df_balanced_nc |>
  filter(model %in% c(REASONING_PAIRS$base, REASONING_PAIRS$reasoning)) |>
  mutate(variant = if_else(model %in% REASONING_PAIRS$reasoning, "reasoning", "base"),
         base_model = if_else(model %in% REASONING_PAIRS$reasoning,
                              str_replace(as.character(model), "-reasoning$", ""),
                              as.character(model))) |>
  select(base_model, variant, all_of(L2_FEATURES)) |>
  pivot_longer(-c(base_model, variant), names_to = "feature", values_to = "value") |>
  filter(!is.na(value)) |>
  group_by(base_model, variant, feature) |>
  summarise(prop_ci(value), .groups = "drop") |>
  mutate(feature = factor(feature, levels = L2_FEATURES))

reasoning_delta <- reasoning_long |>
  select(base_model, variant, feature, rate) |>
  pivot_wider(names_from = variant, values_from = rate) |>
  mutate(delta = reasoning - base)

cross_pair_mean <- reasoning_delta |>
  group_by(feature) |>
  summarise(delta = mean(delta, na.rm = TRUE), .groups = "drop")

ggplot(reasoning_delta, aes(feature, delta, fill = base_model)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.8, alpha = 0.85) +
  geom_point(data = cross_pair_mean, aes(feature, delta),
             inherit.aes = FALSE, shape = 95, size = 12,
             color = "black", alpha = 0.7) +
  geom_hline(yintercept = 0, colour = "grey40") +
  scale_y_continuous(labels = percent_format(1)) +
  labs(x = NULL, y = "Δ (reasoning − base)", fill = "Model pair",
       title = "L2 feature shifts when reasoning is enabled",
       subtitle = "Black bar = mean delta across the four reasoning pairs.") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "bottom")
```

## 6.2 Model family comparison

```{r family-l0}
df_clean |>
  group_by(family) |>
  summarise(
    n_models = n_distinct(model),
    n_records = n(),
    refusal_rate = mean(layer0 == "Non-compliance"),
    partial_rate = mean(layer0 == "Partial compliance"),
    full_rate    = mean(layer0 == "Full compliance"),
    .groups = "drop"
  ) |>
  mutate(across(ends_with("_rate"), ~percent(.x, accuracy = 0.1))) |>
  kable(caption = "Layer 0 distribution by provider family (pooled).") |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

```{r family-l2, fig.height = 6}
df_balanced_nc |>
  select(family, all_of(L2_FEATURES)) |>
  pivot_longer(-family, names_to = "feature", values_to = "value") |>
  filter(!is.na(value)) |>
  group_by(family, feature) |>
  summarise(rate = mean(value), .groups = "drop") |>
  mutate(feature = factor(feature, levels = L2_FEATURES)) |>
  ggplot(aes(feature, family, fill = rate)) +
    geom_tile(colour = "white", linewidth = 0.4) +
    geom_text(aes(label = percent(rate, accuracy = 1)), size = 3, vjust = 1) +
    scale_fill_gradient(low = "#f7f7f7", high = "#08519c", labels = percent_format(1)) +
    labs(x = NULL, y = NULL, fill = "Rate",
         title = "L2 feature rates by provider family (balanced cohort)") +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

### Family rates with Wilson 95% CIs and cross-family mean

Same data as the heatmap above, but presented as a dot-plot with error bars
plus an overlaid cross-family mean. This is the right view for asking "is
there an overall trend for/against a feature, or does it depend on family?"

```{r family-l2-errorbars, fig.height = 6.5}
family_ci <- df_balanced_nc |>
  select(family, all_of(L2_FEATURES)) |>
  pivot_longer(-family, names_to = "feature", values_to = "value") |>
  filter(!is.na(value)) |>
  group_by(family, feature) |>
  summarise(prop_ci(value), .groups = "drop") |>
  mutate(feature = factor(feature, levels = L2_FEATURES))

cross_family_mean <- family_ci |>
  group_by(feature) |>
  summarise(rate = mean(rate), .groups = "drop")

ggplot(family_ci, aes(feature, rate, color = family)) +
  geom_pointrange(aes(ymin = lo, ymax = hi),
                  position = position_dodge(width = 0.6), size = 0.3) +
  geom_point(data = cross_family_mean,
             aes(feature, rate),
             inherit.aes = FALSE, shape = 95, size = 12,
             color = "black", alpha = 0.6) +
  scale_y_continuous(labels = percent_format(1)) +
  labs(x = NULL, y = "Rate within NC (95% Wilson CI)",
       color = "Family",
       title = "L2 feature rates by family — with cross-family mean overlay",
       subtitle = "Black bar = unweighted mean across families.") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "bottom")
```

## 6.3 Flagship-model comparison

A four-way comparison of each provider's flagship: OpenAI (`gpt-5.3`),
Anthropic (`claude-opus-4.6`), Google (`gemini-2.5-pro`), xAI (`grok-4.20`).
Swap the `flagships` vector to use the reasoning variants or older releases.

```{r flagship-setup}
flagships <- c(
  "OpenAI"    = "gpt-5.3",
  "Anthropic" = "claude-opus-4.6",
  "Google"    = "gemini-2.5-pro",
  "xAI"       = "grok-4.20"
)

flag_pal <- c(
  "gpt-5.3"           = "#10a37f",
  "claude-opus-4.6"   = "#d97757",
  "gemini-2.5-pro"    = "#4285f4",
  "grok-4.20"         = "#1d1d1f"
)

df_flag    <- df_balanced |> filter(model %in% flagships)
df_flag_nc <- df_flag    |> filter(layer0 == "Non-compliance")
```

### Layer 0

```{r flagship-l0, fig.height = 4}
df_flag |>
  count(model, layer0) |>
  group_by(model) |>
  mutate(prop = n / sum(n),
         layer0 = factor(layer0, levels = l0_levels),
         model  = factor(model, levels = unname(flagships))) |>
  ggplot(aes(model, prop, fill = layer0)) +
    geom_col(width = 0.7) +
    scale_y_continuous(labels = percent_format(1)) +
    scale_fill_manual(values = l0_pal) +
    labs(x = NULL, y = NULL, fill = "Layer 0",
         title = "Flagship comparison: Layer 0 outcome") +
    theme_minimal(base_size = 12)
```

### Overall L2 distribution

```{r flagship-l2, fig.height = 5}
df_flag_nc |>
  select(model, all_of(L2_FEATURES)) |>
  pivot_longer(-model, names_to = "feature", values_to = "value") |>
  filter(!is.na(value)) |>
  group_by(model, feature) |>
  summarise(rate = mean(value), .groups = "drop") |>
  mutate(feature = factor(feature, levels = L2_FEATURES),
         model   = factor(model, levels = unname(flagships))) |>
  ggplot(aes(feature, rate, fill = model)) +
    geom_col(position = position_dodge(width = 0.9), width = 0.85) +
    scale_y_continuous(labels = percent_format(1), expand = expansion(mult = c(0, 0.05))) +
    scale_fill_manual(values = flag_pal) +
    labs(x = NULL, y = "rate (within NC)", fill = NULL,
         title = "Flagship comparison: L2 feature rates (within Non-compliance)",
         subtitle = "Higher bars indicate a feature appears more often in this model's refusals") +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 30, hjust = 0, vjust = -0.2),
          legend.position = "bottom")
```

### Double-click: L2 by harm category, flagship models only

Heatmap of L2 feature rates faceted by harm category, restricted to the four
flagships. Same `MIN_CELL_N` mask as §5.5 to suppress unreliable rates.

```{r flagship-l2-by-cat, fig.height = 16, fig.width = 10}

flag_l2_cat <- df_flag_nc |>
  select(model, category = primary_llama_guard_category_name,
         all_of(L2_FEATURES)) |>
  pivot_longer(-c(model, category), names_to = "feature", values_to = "value") |>
  filter(!is.na(value)) |>
  group_by(model, category, feature) |>
  summarise(rate = mean(value), n = n(), .groups = "drop") |>
  mutate(rate = if_else(n < MIN_CELL_N, NA_real_, rate),
         feature = factor(feature, levels = L2_FEATURES),
         model   = factor(model, levels = unname(flagships)),
         category = factor(category, levels = cat_order))

ggplot(flag_l2_cat, aes(feature, model, fill = rate)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  geom_text(aes(label = if_else(is.na(rate), "", percent(rate, accuracy = 1))),
            size = 2.6, vjust = 1) +
  facet_wrap(~ category, ncol = 2) +
  scale_fill_gradient(low = "#f7f7f7", high = "#08519c",
                      na.value = "grey92",
                      labels = percent_format(1), name = "Rate") +
  scale_x_discrete(position = "top") +
  labs(x = NULL, y = NULL,
       title = "Flagship × L2 feature × harm category",
       subtitle = paste0("Rate within each (model, category) NC set. Cells with n < ",
                         MIN_CELL_N, " masked.")) +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, hjust = 0, vjust = -0.2),
        strip.text = element_text(face = "bold"))
```

```{r}

flag_l2_cat |>
  ggplot(aes(feature, rate, fill = model)) +
    geom_col(position = position_dodge(width = 0.9), width = 0.85) +
    scale_y_continuous(labels = percent_format(1), expand = expansion(mult = c(0, 0.05))) +
    scale_fill_manual(values = flag_pal) +
    labs(x = NULL, y = "rate (within NC)", fill = NULL,
         title = "Flagship comparison: L2 feature rates (within Non-compliance)",
         subtitle = "Higher bars indicate a feature appears more often in this model's refusals") +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 30, hjust = 0, vjust = -0.2),
          legend.position = "bottom") + 
  facet_grid(category ~ .)

ggsave("harm_type_model.pdf", width = 30, height = 30)

```

### Mitigation profile, side-by-side

```{r flagship-affiliative, fig.height = 4}
df_flag_nc |>
  mutate(model = factor(model, levels = unname(flagships))) |>
  select(model, apology, hedge, positive_alignment, solidarity) |>
  pivot_longer(-model, names_to = "feature", values_to = "value") |>
  filter(!is.na(value)) |>
  group_by(model, feature) |>
  summarise(rate = mean(value), .groups = "drop") |>
  ggplot(aes(feature, rate, fill = model)) +
    geom_col(position = position_dodge(width = 0.9), width = 0.85) +
    scale_y_continuous(labels = percent_format(1), expand = expansion(mult = c(0, 0.05))) +
    scale_fill_manual(values = flag_pal) +
    labs(x = NULL, y = NULL, fill = NULL,
         title = "Flagship affiliative / mitigating features") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
```

## 6.4 Selected hypothesis-shaped views

> *Pre-registered framing direction (paper draft): modern safety training appears
> to shift LLM refusals toward firmer boundaries and away from human-like
> face-saving moves. The plots below set up that contrast — they are descriptive
> only and intentionally avoid significance testing at this stage.*

### Apology decay across OpenAI and Anthropic releases

```{r apology-decay, fig.height = 4}

df_balanced_nc |>
  filter(family %in% c("OpenAI", "Anthropic"),
         !is.na(apology)) |>
  group_by(family, model) |>
  summarise(apology_rate = mean(apology), n = n(), .groups = "drop") |>
  ggplot(aes(model, apology_rate, fill = family)) +
    geom_col() +
    geom_text(aes(label = percent(apology_rate, accuracy = 1)), vjust = -0.4, size = 3) +
    scale_y_continuous(labels = percent_format(1), expand = expansion(mult = c(0, 0.1))) +
    labs(x = NULL, y = "apology rate",
         title = "Apology in non-compliance: OpenAI vs Anthropic",
         subtitle = "Earlier models lean on apology; newer ones do not") +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 30, vjust=0.5, hjust=0.5),
          legend.position = "bottom")

```

### Negative stance is high almost everywhere

```{r negstance-prevalence, fig.height = 4}
df_balanced_nc |>
  filter(!is.na(negative_stance)) |>
  group_by(family, model) |>
  summarise(rate = mean(negative_stance), .groups = "drop") |>
  ggplot(aes(model, rate, fill = family)) +
    geom_col() +
    geom_text(aes(label = percent(rate, accuracy = 1)), vjust = -0.4, size = 3) +
    scale_y_continuous(labels = percent_format(1), expand = expansion(mult = c(0, 0.1))) +
    labs(x = NULL, y = "negative_stance rate",
         title = "Negative stance prevalence by model") +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 30, vjust=0.5, hjust=0.5),
          legend.position = "bottom")
```

### Affiliative features: positive alignment & solidarity

```{r affiliative, fig.height = 5}
df_balanced_nc |>
  select(family, model, positive_alignment, solidarity) |>
  pivot_longer(c(positive_alignment, solidarity), names_to = "feature", values_to = "value") |>
  filter(!is.na(value)) |>
  group_by(family, model, feature) |>
  summarise(rate = mean(value), .groups = "drop") |>
  ggplot(aes(model, rate, fill = feature)) +
    geom_col(position = position_dodge(width = 0.8), width = 0.8) +
    scale_y_continuous(labels = percent_format(1)) +
    labs(x = NULL, y = NULL, fill = NULL,
         title = "Affiliative refusal features by model") +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "bottom")
```

---

# 7. Hard prompts: when one model refuses, what do the others do?

A natural follow-up to the service-level refusal characterisation in §1.2:
if Anthropic blocks a prompt at the API level, **how do OpenAI / Google / Meta /
Qwen etc handle the same prompt in their own voice?** This section subsets to
the prompts that triggered at least one service-level refusal — the hardest
end of the corpus — and asks what *other* models did with them.

## 7.1 The hard-prompt subset

```{r hard-prompts-setup}
HARD_PROMPTS <- df_full |>
  filter(judge_skipped) |>
  pull(final_sample_id) |>
  unique()

df_hard           <- df_full |> filter(final_sample_id %in% HARD_PROMPTS)
df_hard_responses <- df_hard |> filter(!judge_skipped, !parse_error, !is.na(layer0))

# Per-family SLR contribution: how many distinct prompts did each family block?
slr_contrib <- df_full |>
  filter(judge_skipped) |>
  group_by(family) |>
  summarise(n_slr = n(),
            n_distinct_prompts = n_distinct(final_sample_id),
            .groups = "drop")

slr_contrib |>
  kable(caption = paste0("Per-family contribution to the ", length(HARD_PROMPTS),
                         "-prompt hard subset. A prompt counts toward a family if at least one of its models triggered a service-level refusal on it.")) |>
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

## 7.2 L0 distribution on hard prompts, by responding family

Pool across models within each family. Inclusive view: service-level refusals
are visible as the black band on the blocking family's own bar.

```{r hard-l0-by-family, fig.height = 5}
df_hard |>
  mutate(layer0_view = case_when(
    judge_skipped               ~ "Service-Level",
    is.na(layer0)               ~ NA_character_,
    TRUE                         ~ layer0
  )) |>
  filter(!is.na(layer0_view)) |>
  count(family, layer0_view) |>
  group_by(family) |>
  mutate(prop = n / sum(n),
         layer0_view = factor(layer0_view, levels = l0_levels_incl)) |>
  ungroup() |>
  ggplot(aes(family, prop, fill = layer0_view)) +
    geom_col(width = 0.8) +
    scale_y_continuous(labels = percent_format(1)) +
    scale_fill_manual(values = l0_pal_incl) +
    coord_flip() +
    labs(x = NULL, y = NULL, fill = "Layer 0",
         title = "L0 on hard prompts only (inclusive of service-level)",
         subtitle = paste0("Pooled across models within each family. n = ",
                           length(HARD_PROMPTS), " distinct hard prompts.")) +
    theme_minimal(base_size = 11)
```

## 7.3 L1 and L2 shifts on hard prompts vs the overall corpus

Are in-voice refusals on hard prompts *different* from the overall refusal
profile? Compare the pooled NC distribution on the hard subset to the same
distribution on all prompts.

```{r hard-l1-comparison, fig.height = 4}
bind_rows(
  df_clean |> filter(layer0 == "Non-compliance", layer1 %in% l1_levels) |>
    count(layer1) |> mutate(set = "All prompts"),
  df_hard_responses |> filter(layer0 == "Non-compliance", layer1 %in% l1_levels) |>
    count(layer1) |> mutate(set = "Hard prompts only")
) |>
  group_by(set) |>
  mutate(prop = n / sum(n),
         layer1 = factor(layer1, levels = l1_levels)) |>
  ungroup() |>
  ggplot(aes(layer1, prop, fill = set)) +
    geom_col(position = position_dodge(width = 0.8), width = 0.8) +
    scale_y_continuous(labels = percent_format(1)) +
    scale_fill_manual(values = c("All prompts" = "#bdbdbd", "Hard prompts only" = "#7570b3")) +
    labs(x = NULL, y = NULL, fill = NULL,
         title = "L1 distribution shifts on hard prompts",
         subtitle = "In-voice refusals only. Hard prompts tend to shift L1 toward Ethics-based and Bare.") +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom")
```

```{r hard-l2-comparison, fig.height = 5}
bind_rows(
  df_balanced_nc |>
    select(all_of(L2_FEATURES)) |>
    pivot_longer(everything(), names_to = "feature", values_to = "value") |>
    filter(!is.na(value)) |>
    group_by(feature) |>
    summarise(prop_ci(value), .groups = "drop") |>
    mutate(set = "Balanced cohort (baseline)"),
  df_hard_responses |>
    filter(layer0 == "Non-compliance") |>
    select(all_of(L2_FEATURES)) |>
    pivot_longer(everything(), names_to = "feature", values_to = "value") |>
    filter(!is.na(value)) |>
    group_by(feature) |>
    summarise(prop_ci(value), .groups = "drop") |>
    mutate(set = "Hard prompts only")
) |>
  mutate(feature = factor(feature, levels = L2_FEATURES)) |>
  ggplot(aes(feature, rate, color = set)) +
    geom_pointrange(aes(ymin = lo, ymax = hi),
                    position = position_dodge(width = 0.5), size = 0.3) +
    scale_y_continuous(labels = percent_format(1)) +
    scale_color_manual(values = c("Balanced cohort (baseline)" = "#bdbdbd",
                                  "Hard prompts only" = "#7570b3")) +
    labs(x = NULL, y = "Rate (95% Wilson CI)", color = NULL,
         title = "L2 form features: hard prompts vs balanced cohort",
         subtitle = "Within in-voice refusals. Diverging dots = form of refusal hardens (or softens) on hard prompts.") +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "bottom")
```

## 7.4 Per-blocker conditional: what did the other families do?

For each blocking family B, take the prompts B blocked at the API level, then
look at the L0 distribution of every *other* family's response to those same
prompts. Diagonal cells (blocker = responder) include B's own non-blocked
responses to its own blocked prompts (e.g., a different Anthropic model that
did respond).

```{r blocker-responder-l0, fig.height = 6}
# For each prompt, the set of families that produced a service-level refusal
prompts_by_blocker <- df_full |>
  filter(judge_skipped) |>
  select(family_blocker = family, final_sample_id) |>
  distinct()

# Join each blocker prompt with every in-voice response (any model) to that prompt
blocker_responder <- prompts_by_blocker |>
  inner_join(
    df_hard_responses |> rename(family_responder = family),
    by = "final_sample_id",
    relationship = "many-to-many"
  )

blocker_responder |>
  count(family_blocker, family_responder, layer0) |>
  group_by(family_blocker, family_responder) |>
  mutate(prop = n / sum(n),
         layer0 = factor(layer0, levels = l0_levels)) |>
  ungroup() |>
  ggplot(aes(family_responder, prop, fill = layer0)) +
    geom_col(width = 0.85) +
    facet_wrap(~ family_blocker, ncol = 4,
               labeller = labeller(family_blocker = function(x) paste0("Blocked by ", x))) +
    scale_y_continuous(labels = percent_format(1)) +
    scale_fill_manual(values = l0_pal) +
    coord_flip() +
    labs(x = NULL, y = NULL, fill = "Layer 0",
         title = "On prompts blocker family refused at API: what did the responder family do?",
         subtitle = "Each panel = one blocking family. Bars = L0 distribution of in-voice responses from each responder.") +
    theme_minimal(base_size = 10) +
    theme(legend.position = "bottom",
          strip.text = element_text(face = "bold"))
```

```{r blocker-responder-nc-rate}
blocker_responder |>
  group_by(family_blocker, family_responder) |>
  summarise(
    n_responses = n(),
    nc_rate     = mean(layer0 == "Non-compliance"),
    full_rate   = mean(layer0 == "Full compliance"),
    .groups = "drop"
  ) |>
  mutate(nc_rate   = percent(nc_rate, accuracy = 0.1),
         full_rate = percent(full_rate, accuracy = 0.1)) |>
  arrange(family_blocker, desc(nc_rate)) |>
  kable(caption = "On prompts each blocker family API-refused, what proportion of responders' responses were NC vs Full compliance?") |>
  kable_styling(bootstrap_options = c("striped","hover"), font_size = 10)
```

## 7.5 Per-blocker L2 form features

For each blocker family, the L2 profile of in-voice refusals from each
responder. Use only the most diagnostic features so the figure stays readable.

```{r blocker-responder-l2, fig.height = 7}
KEY_L2 <- c("explicit_nc", "apology", "negative_stance", "alternative_offer",
            "executed_alternative", "statement_of_principle", "solidarity")

blocker_responder |>
  filter(layer0 == "Non-compliance") |>
  select(family_blocker, family_responder, all_of(KEY_L2)) |>
  pivot_longer(all_of(KEY_L2), names_to = "feature", values_to = "value") |>
  filter(!is.na(value)) |>
  group_by(family_blocker, family_responder, feature) |>
  summarise(rate = mean(value), n = n(), .groups = "drop") |>
  filter(n >= 5) |>
  mutate(feature = factor(feature, levels = KEY_L2)) |>
  ggplot(aes(feature, family_responder, fill = rate)) +
    geom_tile(colour = "white", linewidth = 0.3) +
    geom_text(aes(label = percent(rate, accuracy = 1)), size = 2.6, vjust = 1) +
    facet_wrap(~ family_blocker, ncol = 4,
               labeller = labeller(family_blocker = function(x) paste0("Blocked by ", x))) +
    scale_fill_gradient(low = "#f7f7f7", high = "#08519c",
                        labels = percent_format(1), name = "Rate") +
    scale_x_discrete(position = "top") +
    labs(x = NULL, y = NULL,
         title = "L2 form features of in-voice refusals on hard prompts",
         subtitle = "Each panel = blocker family. Rows = responder family. Cells = feature rate among NC responses (n ≥ 5 required).") +
    theme_minimal(base_size = 9) +
    theme(axis.text.x = element_text(angle = 30, hjust = 0, vjust = -0.2),
          strip.text = element_text(face = "bold"))
```

## 7.6 Headline cross-provider stats

A single-line summary highlighting the "Anthropic blocks → what does OpenAI do"
question that motivated this section.

```{r hard-headline}
target_pairs <- tribble(
  ~family_blocker, ~family_responder,
  "Anthropic",     "OpenAI",
  "Anthropic",     "Meta",
  "Anthropic",     "Qwen",
  "Google",        "OpenAI",
  "Google",        "Anthropic",
  "xAI",           "OpenAI",
  "xAI",           "Anthropic"
)

target_pairs |>
  left_join(
    blocker_responder |>
      mutate(family_blocker   = as.character(family_blocker),
             family_responder = as.character(family_responder)) |>
      group_by(family_blocker, family_responder) |>
      summarise(
        n = n(),
        nc_rate     = mean(layer0 == "Non-compliance"),
        full_rate   = mean(layer0 == "Full compliance"),
        bare_rate   = mean(layer1 == "Bare", na.rm = TRUE),
        ethics_rate = mean(layer1 == "Ethics-based", na.rm = TRUE),
        apology_rate = mean(apology, na.rm = TRUE),
        .groups = "drop"
      ),
    by = c("family_blocker", "family_responder")
  ) |>
  mutate(across(c(nc_rate, full_rate, bare_rate, ethics_rate, apology_rate),
                ~ifelse(is.na(.x), NA, percent(.x, accuracy = 0.1)))) |>
  kable(caption = "Headline cross-provider stats: when blocker family refused at API, how did responder family handle the same prompt?") |>
  kable_styling(bootstrap_options = c("striped","hover"), font_size = 10)
```

---

# 8. Flavour: sampled responses per model

This section samples a small number of responses per (model × Layer 0 outcome)
so the reader can ground numerical findings in concrete language. Responses are
**not cherry-picked**: they are drawn at random with a fixed seed inside this
chunk. Long responses are truncated to keep the document readable.

> **Note.** This section will surface the raw user queries and model responses,
> which include harmful prompts from SorryBench/LMSYS. View with awareness; the
> code generates the sample without anyone needing to read the corpus first.

```{r flavour-config}
set.seed(20260519)
MAX_QUERY_CHARS    <- 400
MAX_RESPONSE_CHARS <- 900
SAMPLES_PER_CELL   <- 1   # per (model × L0). Bump to 2-3 once layout is reviewed.

# HTML-escape user-facing text. Also substitutes entities for the two
# characters that can launch block-level markdown constructs (#, `), as a
# defence in depth in case pandoc's markdown_in_html_blocks extension
# parses content inside our table cells.
html_escape <- function(x) {
  x <- as.character(x)
  x <- gsub("&", "&amp;", x, fixed = TRUE)
  x <- gsub("<", "&lt;",  x, fixed = TRUE)
  x <- gsub(">", "&gt;",  x, fixed = TRUE)
  x <- gsub("#", "&#35;", x, fixed = TRUE)
  x <- gsub("`", "&#96;", x, fixed = TRUE)
  x
}
truncate_text <- function(x, max_chars) {
  x <- as.character(x)
  ifelse(nchar(x) > max_chars,
         paste0(substr(x, 1, max_chars - 3), "..."),
         x)
}

# Render a tibble of flavour rows as a raw HTML table. Pandoc preserves
# <table> blocks verbatim with the raw_html extension, and content inside
# <pre>...</pre> is not parsed as markdown — together this avoids kable's
# pipe-format / escape-on-render quirks that were leaking model markdown
# into the document's ToC.
render_flavour_table <- function(df, text_cols = c("query", "response"),
                                 col_widths = list(query = "22em",
                                                   response = "32em")) {
  meta_cols <- setdiff(names(df), text_cols)

  css <- '<style>
    table.flavour { border-collapse: collapse; width: 100%; font-size: 11px;
                    font-family: ui-sans-serif, system-ui, sans-serif; }
    table.flavour th, table.flavour td { border: 1px solid #ddd; padding: 6px;
                                         vertical-align: top; }
    table.flavour th { background: #f4f4f4; font-weight: 600; text-align: left; }
    table.flavour tr:nth-child(even) td { background: #fafafa; }
    table.flavour pre { white-space: pre-wrap; margin: 0;
                        font-family: ui-monospace, monospace; font-size: 11px;
                        max-height: 18em; overflow-y: auto;
                        background: transparent; border: none; padding: 0; }
  </style>'

  thead <- paste0("<tr>",
                  paste0("<th>", html_escape(names(df)), "</th>", collapse = ""),
                  "</tr>")

  body_rows <- vapply(seq_len(nrow(df)), function(i) {
    cells <- vapply(names(df), function(col) {
      val <- df[[col]][[i]]
      if (col %in% text_cols) {
        width <- col_widths[[col]] %||% "20em"
        paste0('<td style="max-width:', width, ';"><pre>',
               html_escape(truncate_text(val,
                                         if (col == "response") MAX_RESPONSE_CHARS
                                         else MAX_QUERY_CHARS)),
               "</pre></td>")
      } else {
        paste0("<td>", html_escape(val), "</td>")
      }
    }, character(1))
    paste0("<tr>", paste0(cells, collapse = ""), "</tr>")
  }, character(1))

  paste0(css, '<table class="flavour"><thead>', thead,
         "</thead><tbody>", paste0(body_rows, collapse = ""),
         "</tbody></table>")
}

`%||%` <- function(a, b) if (is.null(a)) b else a
```

## 7.1 One example per (model × L0)

```{r flavour-by-model, results = 'asis'}
flavour <- df_clean |>
  filter(!is.na(query), !is.na(response), nchar(response) > 0) |>
  group_by(model, layer0) |>
  slice_sample(n = SAMPLES_PER_CELL) |>
  ungroup() |>
  arrange(model, factor(layer0, levels = l0_levels)) |>
  mutate(
    L2 = pmap_chr(
      pick(all_of(L2_FEATURES)),
      function(...) {
        vals <- list(...)
        names(vals) <- L2_FEATURES
        on <- names(vals)[vapply(vals, function(v) isTRUE(as.integer(v) == 1L), logical(1))]
        if (length(on) == 0) "—" else paste(on, collapse = ", ")
      }
    )
  ) |>
  select(model, layer0, layer1, L2,
         category = primary_llama_guard_category_name,
         query, response)

cat("<p><i>One sampled response per (model × Layer 0 outcome).</i></p>\n")
cat(render_flavour_table(flavour))
```

## 7.2 One example per (harm category × representative model)

To get a flavour across the request spectrum: for each harm category, sample
one response from `gpt-4o`, `claude-sonnet-4.6`, `gemini-2.5-pro`, and
`llama-3.1-70b` (a spread of provider safety stances).

```{r flavour-by-category, results = 'asis'}
focal_models <- c("gpt-4o", "claude-sonnet-4.6", "gemini-2.5-pro", "llama-3.1-70b")

flavour_by_cat <- df_clean |>
  filter(model %in% focal_models,
         !is.na(query), !is.na(response), nchar(response) > 0) |>
  group_by(primary_llama_guard_category_name, model) |>
  slice_sample(n = 1) |>
  ungroup() |>
  arrange(primary_llama_guard_category_name,
          factor(model, levels = focal_models)) |>
  select(category = primary_llama_guard_category_name,
         model, layer0, layer1, query, response)

cat("<p><i>One sampled response per harm category × focal model.</i></p>\n")
cat(render_flavour_table(flavour_by_cat))
```

---

# 9. Statistical hypothesis tests (scaffolding)

>  The blocks
> below give a working starting point — chi-square omnibus tests, McNemar for
> the paired reasoning comparisons, a mixed-effects model that respects the
> nested structure (prompts × models), and a composite "human-like mitigation"
> score. Adapt freely; report what survives multiple-comparison correction.

```{r stats-setup}
# Optional packages; the chunk degrades gracefully if missing.
has_lme4   <- requireNamespace("lme4",   quietly = TRUE)
has_broomm <- requireNamespace("broom.mixed", quietly = TRUE)
if (has_lme4)   suppressMessages(library(lme4))
if (has_broomm) suppressMessages(library(broom.mixed))

# Prompt identifier shared across models (same 200 prompts hit every model).
df_clean <- df_clean |>
  mutate(prompt_id = coalesce(final_sample_id, candidate_id))
df_nc <- df_nc |>
  mutate(prompt_id = coalesce(final_sample_id, candidate_id))
```

## 8.1 Omnibus: do models differ?

```{r omnibus-model-l0}
tab_model_l0 <- df_clean |>
  count(model, layer0) |>
  pivot_wider(names_from = layer0, values_from = n, values_fill = 0) |>
  column_to_rownames("model") |>
  as.matrix()

chi_model <- chisq.test(tab_model_l0)
cramers_v <- sqrt(chi_model$statistic /
                  (sum(tab_model_l0) * (min(dim(tab_model_l0)) - 1)))

tibble(
  test = "χ²(L0 × model)",
  statistic = round(chi_model$statistic, 2),
  df = chi_model$parameter,
  p_value = format.pval(chi_model$p.value, digits = 3),
  cramers_v = round(as.numeric(cramers_v), 3)
) |>
  kable(caption = "Models differ in Layer 0 outcome (omnibus). Cramér's V quantifies the effect size; >0.1 small, >0.3 medium, >0.5 large.") |>
  kable_styling(full_width = FALSE)
```

## 8.2 Omnibus: do harm categories elicit different actions?

```{r omnibus-cat-l0}
tab_cat_l0 <- df_clean |>
  count(primary_llama_guard_category_name, layer0) |>
  pivot_wider(names_from = layer0, values_from = n, values_fill = 0) |>
  column_to_rownames("primary_llama_guard_category_name") |>
  as.matrix()

chi_cat <- chisq.test(tab_cat_l0)
v_cat   <- sqrt(chi_cat$statistic /
                (sum(tab_cat_l0) * (min(dim(tab_cat_l0)) - 1)))

tibble(
  test = "χ²(L0 × harm category)",
  statistic = round(chi_cat$statistic, 2),
  df = chi_cat$parameter,
  p_value = format.pval(chi_cat$p.value, digits = 3),
  cramers_v = round(as.numeric(v_cat), 3)
) |>
  kable(caption = "Harm categories differ in elicited Layer 0 outcome (pooled across models).") |>
  kable_styling(full_width = FALSE)
```

## 8.3 Reasoning vs base: McNemar paired tests

The same 200 prompts are sent to both the base and reasoning variant of a
model, so the right test for "does turning on reasoning change refusal
behaviour?" is paired. McNemar's test on the discordant cells of the 2×2
table is the standard choice for a binary outcome.

```{r mcnemar-l0}
mcnemar_one <- function(base_model, reasoning_model, outcome_col, outcome_value) {
  pair <- df_clean |>
    filter(model %in% c(base_model, reasoning_model)) |>
    mutate(variant = if_else(model == reasoning_model, "reasoning", "base"),
           hit = .data[[outcome_col]] == outcome_value) |>
    select(prompt_id, variant, hit) |>
    pivot_wider(names_from = variant, values_from = hit, values_fn = first) |>
    drop_na(base, reasoning)

  tab <- table(base = pair$base, reasoning = pair$reasoning)
  if (any(dim(tab) < 2)) return(tibble(test = NA, statistic = NA, p_value = NA, n = nrow(pair)))
  m <- mcnemar.test(tab)
  tibble(statistic = unname(m$statistic), p_value = m$p.value, n = nrow(pair))
}

l0_pairs <- REASONING_PAIRS |>
  rowwise() |>
  mutate(res = list(mcnemar_one(base, reasoning, "layer0", "Non-compliance"))) |>
  unnest(res) |>
  ungroup() |>
  mutate(p_bh = p.adjust(p_value, "BH"))

l0_pairs |>
  mutate(p_value = format.pval(p_value, digits = 3),
         p_bh    = format.pval(p_bh,    digits = 3),
         statistic = round(statistic, 2)) |>
  kable(caption = "McNemar test on `refused (NC) vs not` for each base/reasoning pair.") |>
  kable_styling(full_width = FALSE)
```

```{r mcnemar-l2}
# Same logic, applied to every L2 feature × pair.
mcnemar_feature <- function(base_model, reasoning_model, feature) {
  pair <- df_clean |>
    filter(model %in% c(base_model, reasoning_model),
           !is.na(.data[[feature]])) |>
    mutate(variant = if_else(model == reasoning_model, "reasoning", "base"),
           hit = .data[[feature]] == 1) |>
    select(prompt_id, variant, hit) |>
    pivot_wider(names_from = variant, values_from = hit, values_fn = first) |>
    drop_na(base, reasoning)

  tab <- table(base = pair$base, reasoning = pair$reasoning)
  if (any(dim(tab) < 2) || nrow(pair) < 10)
    return(tibble(statistic = NA_real_, p_value = NA_real_, n = nrow(pair)))
  m <- mcnemar.test(tab)
  tibble(statistic = unname(m$statistic), p_value = m$p.value, n = nrow(pair))
}

l2_pairs <- REASONING_PAIRS |>
  crossing(feature = L2_FEATURES) |>
  rowwise() |>
  mutate(res = list(mcnemar_feature(base, reasoning, feature))) |>
  unnest(res) |>
  ungroup() |>
  group_by(base) |>
  mutate(p_bh = p.adjust(p_value, "BH")) |>
  ungroup()

l2_pairs |>
  filter(!is.na(p_value)) |>
  mutate(p_value = format.pval(p_value, digits = 3),
         p_bh    = format.pval(p_bh,    digits = 3),
         statistic = round(statistic, 2)) |>
  arrange(base, factor(feature, levels = L2_FEATURES)) |>
  kable(caption = "L2 feature shifts under reasoning, paired McNemar with BH-corrected p-values within each model pair.") |>
  kable_styling(full_width = FALSE, font_size = 10)
```

## 8.4 Mixed-effects model: model effect on refusal, controlling for prompt

A chi-square treats every (model × prompt) cell as independent, which they
aren't — the same 200 prompts hit all 16 models. A logistic mixed model with
a random intercept per prompt is the cleanest "model X refuses more than
model Y, controlling for prompt difficulty" estimator.

```{r glmm-refusal, eval = has_lme4}
fit_data <- df_clean |>
  mutate(refused = as.integer(layer0 == "Non-compliance"),
         model   = factor(model, levels = MODEL_ORDER),
         prompt_id = factor(prompt_id)) |>
  filter(!is.na(refused), !is.na(prompt_id))

# Reference level: gpt-4o (oldest and most permissive OpenAI in the set).
fit_data$model <- relevel(fit_data$model, ref = "gpt-4o")

m_refuse <- glmer(refused ~ model + (1 | prompt_id),
                  data = fit_data,
                  family = binomial,
                  control = glmerControl(optimizer = "bobyqa"))

if (has_broomm) {
  broom.mixed::tidy(m_refuse, effects = "fixed", conf.int = TRUE, exponentiate = TRUE) |>
    filter(term != "(Intercept)") |>
    mutate(term = str_replace(term, "^model", ""),
           across(c(estimate, conf.low, conf.high), ~round(.x, 2)),
           p.value = format.pval(p.value, digits = 3)) |>
    select(model = term, OR = estimate, ci_lo = conf.low, ci_hi = conf.high, p = p.value) |>
    kable(caption = "Odds ratios for refusal vs gpt-4o, from glmer(refused ~ model + (1|prompt)). OR > 1 means the model refuses more than gpt-4o on a given prompt.") |>
    kable_styling(full_width = FALSE, font_size = 11)
} else {
  print(summary(m_refuse))
}
```

```{r glmm-refusal-fallback, eval = !has_lme4, echo = FALSE, results = 'asis'}
cat("> `lme4` not installed; mixed-effects table skipped. Install with `install.packages('lme4')` to enable.\n")
```

## 8.5 Composite "human-like mitigation" score

Pre-registered framing: modern safety training shifts refusals away from
human-like face-saving moves. Operationalise a per-response score as the
count of affiliative + mitigating features present (range 0–4):

```{r composite-score}
mitigation_features <- c("apology", "hedge", "positive_alignment", "solidarity")

df_nc <- df_nc |>
  mutate(mitigation_score = rowSums(across(all_of(mitigation_features)), na.rm = TRUE))

df_nc |>
  group_by(family, model) |>
  summarise(n = n(),
            mean_score = mean(mitigation_score, na.rm = TRUE),
            sd_score   = sd(mitigation_score, na.rm = TRUE),
            .groups = "drop") |>
  arrange(desc(mean_score)) |>
  mutate(across(c(mean_score, sd_score), ~round(.x, 2))) |>
  kable(caption = "Mean count of affiliative/mitigating features per NC response (max 4).") |>
  kable_styling(full_width = FALSE, font_size = 11)
```

```{r composite-by-family}
# One-way comparison of mitigation_score across families.
kw_family <- kruskal.test(mitigation_score ~ family, data = df_nc)
tibble(
  test = "Kruskal–Wallis: mitigation_score ~ family",
  statistic = round(kw_family$statistic, 2),
  df = kw_family$parameter,
  p_value = format.pval(kw_family$p.value, digits = 3)
) |>
  kable() |>
  kable_styling(full_width = FALSE)
```

```{r composite-mixed, eval = has_lme4}
# Mixed model on the composite: family effect with prompt random intercept.
fit_score <- df_nc |>
  mutate(prompt_id = factor(prompt_id),
         family    = factor(family))

m_score <- lmer(mitigation_score ~ family + (1 | prompt_id), data = fit_score)
if (has_broomm) {
  broom.mixed::tidy(m_score, effects = "fixed", conf.int = TRUE) |>
    mutate(term = str_replace(term, "^family", ""),
           across(c(estimate, conf.low, conf.high), ~round(.x, 2))) |>
    kable(caption = "Mixed model: mitigation_score ~ family + (1|prompt). Reference family = OpenAI.") |>
    kable_styling(full_width = FALSE)
} else {
  print(summary(m_score))
}
```

## 8.6 Caveats to flag in any reported test

- **Independence.** The same 200 prompts hit every model; pooled chi-squares
  understate uncertainty. Where models are the unit of analysis, prefer the
  mixed-effects estimates in 8.4 / 8.5.
- **Judge as a fixed effect.** All results are conditional on GPT-5.5's
  judgements at this prompt. Sensitivity analysis on a different judge model
  is owed before any test result becomes a paper claim.
- **Multiple comparisons.** L2 has 13 binary features × 4 reasoning pairs =
  52 paired tests; the section already applies BH within each pair. For
  cross-model L2 comparisons (one feature, 16 models, 120 pairwise tests),
  prefer omnibus + a few principled contrasts over exhaustive pairwise.
- **Provider safety blocks.** Gemini's 19 Google-blocked prompts and Grok's
  19 xAI-blocked prompts are excluded from `df_clean` here. If you want to
  count them as system-level NC, recompute with those rows added back —
  they will shift refusal rates and L1 distributions for those two models.

---

# Appendix: Data sources

- Judged data: `data/annotations/responses_<model>.judged.csv`, 16 model files.
- Judge: `openai/gpt-5.5` non-reasoning, prompt SHA256 prefix
  `r substr(PROMPT_SHA256, 1, 12)`...
- Codebook: `taxonomy_guideline (11).pdf`.
- Prompts: 200 sampled from LMSYS/WildChat (`source = "lmsys"`) and
  SorryBench (`source = "sorrybench"`), balanced across Llama Guard harm
  categories.
- Skips (provider-blocked / refusal-stop / transient) and the run-1 OR-credit
  parse_errors are documented in `logs/` and excluded from the analyses above.
